---
title: "Boxplots with Extras"
subtitle: "Enhancing basic boxplots"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document:
    highlight: kate
---

First we load the data (an extension of the dataset described by Rate & McGrath, 
2022) and the R packages we need:

```{r load-data-and-packages, message=FALSE, warning=FALSE, results='hold'}
git <- "https://github.com/Ratey-AtUWA/Learn-R-web/raw/refs/heads/main/"
afs1923 <- read.csv(paste0(git,"afs1923.csv"), stringsAsFactors = TRUE)

library(viridis)        # accessible colour palettes
library(rcompanion)     # utilities for pairwise comparisons
library(multcompView)   # utilities for pairwise comparisons
library(TeachingDemos)  # used only for the shadowtext() function
```

## Boxplots with clear comparisons and intelligible scales

For exploratory data analysis it's common to draw a lot of boxplots. The data we
often explore in Environmental Science often includes variables having
positively skewed distributions. When drawing boxplots, this means that some
categories can be hard to visualise because they have much lower concentrations
than others, and the high values dominate the plot (see Figure 1(a)).

We commonly deal with this by log transformation - Figure 1(b) shows this, and
the variables (and therefore boxplot boxes) now have more symmetrical
distributions which are all clearly visible on the plot. But the *y* axis
(concentration, in this case) doesn't show the actual values.

Figure 1(c) shows the plot with a log-transformed *y* axis scale, but this is
not really satisfying either since the distribution is still skewed but plotted
on the compressed log-transformed *y*-axis.

I think the best option, shown in Figure 1(d), is to plot the log-transformed
variable, but suppress the automatic axis and manually add a *y*-axis with a
transformed scale. It's a bit trickier, but worth it. Now we can compare all the
factor levels easily, yet still have actual values on the axis (*e.g*. for
comparison with environmental guidelines).

```{r read data, include=FALSE, echo=FALSE}
afs1923 <- read.csv(file="afs1923.csv", 
                      stringsAsFactors = TRUE)
```

```{r boxplots-2x2, fig.height=7, fig.width=7, fig.cap="Figure 1: Four styles of boxplots based on the same data.", results='hold'}
par(mfrow=c(2,2),mar = c(5,4,1.5,1), mgp=c(2.4,0.2,0), tcl=0.25, font.lab = 2,
    xpd = F)
library(viridis)

with(afs1923, boxplot(Zn ~ Zone, col=viridis(12, alpha=0.5), 
                        las=2, xlab="", cex = 1.4, cex.lab = 1.4))
mtext("Sampling Zone",1,3,font=2)
mtext("(a)", 3, -1.2, adj = 0.03, font = 2)
mtext("Untransformed variable or axis", 3, 0.2, font=3)

with(afs1923, boxplot(log10(Zn) ~ Zone, col=viridis(12, alpha=0.5), 
                        las=2, xlab="", cex = 1.4, cex.lab = 1.4,
                        ylab=expression(bold(paste(log[10],"(Zn)")))))
mtext("Sampling Zone",1,3,font=2)
mtext("(b)", 3, -1.2, adj = 0.03, font = 2)
mtext("Log-transformed variable", 3, 0.2, font=3)

with(afs1923, boxplot(Zn ~ Zone, log = "y", col=viridis(12, alpha=0.5), 
                        las=2, xlab="", cex = 1.4, cex.lab = 1.4))
mtext("Sampling Zone",1,3,font=2)
mtext("(c)", 3, -1.2, adj = 0.03, font = 2)
mtext(expression(paste(italic("Log-transformed variable "),bold("axis"))), 
      3, 0.2, font=3)

with(afs1923, boxplot(log10(Zn) ~ Zone, col=viridis(12, alpha=0.5), 
                        las=2, xlab="", cex = 1.4, cex.lab = 1.4, yaxt="n",
                        ylab = "Zn"))
logtx <- c(0.1,0.2,0.5,1,2,5,10,20,50,100,100,200,500,1000,2000,5000,10000)
axis(2, at=log10(logtx), labels=logtx, las=2)
mtext("Sampling Zone",1,3,font=2)
mtext("(d)", 3, -1.2, adj = 0.03, font = 2, col = "blue3")
abline(h=par("usr")[4], col="gold", lwd=9)
mtext("Best option: log variable, custom axis", 3, 0.2, font=3, col = "blue3")
```

&nbsp;

## Boxplots with added means or observations

Of course, this is not the only modification we can make with box plots. A box 
plot is a good way to compare the spread of values between different levels of a
factor, and the thick line inside boxes shows how the medians in each group
vary. Sometimes we also like to look at where the mean values are, especially if
we have a parametric statistical test to compare these means. It's not too
difficult to superimpose the means together with some measure of variability;
the following example (Figure 2(a)) shows 95% confidence intervals around each
group mean. We use the `tapply()` function to make columns within a new data
frame which calculate confidence intervals and mean for a variable in groups
defined by a factor. The confidence are available from the output of the 
`t.test()` function, so we use simple custom `function` statements within 
`tapply()` to generate just the confidence interval output.

```{r boxplots-with-means, fig.height=4.5, fig.width=9, fig.cap="Figure 2: Including additional information on boxplots; (a) mean and CI, (b) individual observations, with the sediment GV-high concentration for Zn (410 mg/kg; Water Quality Australia 2024).", results='hold'}
par(mfrow=c(1,2), mar=c(6,4,1,1), mgp=c(2.4,0.2,0), tcl=0.25, font.lab = 2, las=1)
# change data object, variable & factor names to suit your data 
ci0 <- data.frame(lo95=with(afs1923, 
                      tapply(log10(Zn), Zone, function(x){t.test(x)$conf.int[1]})),
                  mean=with(afs1923, 
                      tapply(log10(Zn), Zone, function(x){mean(x, na.rm=TRUE)})),
                  hi95=with(afs1923, 
                      tapply(log10(Zn), Zone, function(x){t.test(x)$conf.int[2]})))
with(afs1923, 
     boxplot(log10(Zn) ~ Zone, col=viridis(nrow(ci0), alpha=0.5),
             ylab="Zn (mg/kg)",
             cex.lab=1.2, yaxt="n", xaxt = "n", xlab = "")
) 
axis(1, las = 2, tcl = 0.4, at = seq(1,NROW(ci0)),
     labels = as.character(levels(afs1923$Zone)))
axis(2, at=log10(logtx), labels=logtx, las=2)
mtext("Sampling zone", side = 1, line = 3.5, font = 2, cex = 1.2)

# use arrow function to make background for [optional] error bars
arrows(x0=seq(1,nrow(ci0)), y0=ci0[,2], y1=ci0[,1], 
       col="white", angle=90, length=0.1, lwd=6) # optional
arrows(x0=seq(1,nrow(ci0)), y0=ci0[,2], y1=ci0[,3], 
       col="white", angle=90, length=0.1, lwd=6) # optional
#
# draw lines to join points first, so the points overplot lines
lines(seq(1,NROW(ci0)), ci0[,2], col="white", lwd=3, type="c")
lines(seq(1,NROW(ci0)), ci0[,2], col="gray40", lwd=1, type="c", lty=3)
#
# use arrow function to make actual [optional] error bars
arrows(x0=seq(1,nrow(ci0)), y0=ci0[,2], y1=ci0[,1], 
       col="gray40", angle=90, length=0.1, lwd=2, lend=1)
arrows(x0=seq(1,nrow(ci0)), y0=ci0[,2], y1=ci0[,3], 
       col="gray40", angle=90, length=0.1, lwd=2, lend=1)
# draw the points with white background for contrast
points(seq(1,NROW(ci0)), ci0[,2], col="white", pch=16, lwd=2, cex=1.6)
points(seq(1,NROW(ci0)), ci0[,2], col="gray40", 
       pch=16, lwd=2, cex=1.2)
mtext("(a)", 3, -1.2, adj = 0.03)

nc <- nlevels(afs1923$Type)
palette(c("black",viridis(nc, alpha=0.5),
          viridis(nlevels(afs1923$Type), alpha=0.75),"white"))
with(afs1923, 
     boxplot(log10(Zn) ~ Type, col=c(1:nc)+1,
             ylab="Zn (mg/kg)",
             cex.lab=1.2, yaxt="n", xlab = "",las=2)) 
axis(2, at=log10(logtx), labels=logtx, las=2)
mtext("Sample Type", 1, 4.5, cex=1.2, font=2)
abline(h=log10(410), col="dimgray", lty="22", lwd=2)
with(afs1923, stripchart(log10(Zn) ~ Type, add = TRUE, vertical=TRUE, 
                           method="jitter", pch=19,col=c(1:nc)+(nc+1)))
mtext("(b)", 3, -1.2, adj = 0.97)
legend("bottomleft", 
       legend="Upper guideline value for\nZn in  sediment (410 mg/kg)",
       bty="n", inset=c(0,0.05), lwd=2, lty="22", seg.len=3, col="dimgray")
```

&nbsp;

Figure 2(b) shows the actual observations, and an environmental threshold for 
Zn. Showing the individual values is probably more useful when there are fewer observations, in which case the boxplots themselves may not represent the distributions well. In addition, including the individual values with an 
environmental threshold means we can [approximately] count the number of samples 
exceeding the threshold.

## Boxplots with pairwise comparisons

We often use a boxplot to visualise a statistical comparison of means 
(including non-parametric comparisons such as a Kruskal-Wallis test). If the 
overall test shows a significant effect (*i.e*. p&nbsp;&le;&nbsp;0.05), we
follow up with a *post-hoc* pairwise comparison such as a pairwise t-test or
Wilcoxon test. The significant pairwise differences are conveniently and
concisely summarised in the “compact letter display”, in which factor levels
having no letters in common have significant pairwise differences.

The compact letters are often presented as superscripts or suffixes in a table
of means, but they can also be included on box plots, as shown in the code
below.

```{r boxplot-compact-letters, fig.height=4.5, fig.width=9, fig.cap="Figure 3: Including pairwise compact letters on boxplots; (a) compact letters at mean values; (b) compact letters at selected coordinates. Categories having no letters in common have significant pairwise differences (p ≤ 0.05).", results='hold'}
par(mfrow=c(1,2), mar=c(6,4,1,1), mgp=c(2.4,0.2,0), tcl=0.25, font.lab = 2, 
    las=2)
# 1. compact letters at mean values for each level
with(afs1923, boxplot(log10(Zn) ~ Type, yaxt="n",             # suppress axes
                    col=viridis(nlevels(Type), alp=0.4, end=0.9),
                    xlab="", ylab="Zn (mg/kg)"))
mtext("Sample Type", 1, 4.5, cex=1, font=2, las=1)
# manual log-transf axis
axis(2, at=log10(logtx), labels=logtx, mgp=c(1.6,0.2,0), 
     tcl=0.25)
meansZn <- with(afs1923, 
                tapply(log10(Zn), Type, function(x){mean(x, na.rm=TRUE)}))
pwZn <- with(afs1923, pairwise.wilcox.test(Zn, Type))
mclZn <- multcompLetters(fullPTable(pwZn$p.value)) 
# plot compact letters at mean values
shadowtext(c(1:4), meansZn, labels=mclZn$Letters, col=1, bg=10, r=0.2)

# 2. compact letters near axis
rgZn <- range(log10(afs1923$Zn), na.rm=TRUE)
with(afs1923, boxplot(log10(Zn) ~ Type, yaxt="n",             # suppress axes
                    col=viridis(nlevels(Type), alp=0.4, end=0.9),
                    xlab="", ylab="Zn (mg/kg)",
                    ylim=c(rgZn[1]-0.7,rgZn[2])))
mtext("Sample Type", 1, 4.5, cex=1, font=2, las=1)
# manual log-transf axis
axis(2, at=log10(logtx[-c(1:3)]), labels=logtx[-c(1:3)], mgp=c(1.6,0.2,0), 
     tcl=0.25)
pwZn <- with(afs1923, pairwise.wilcox.test(Zn, Type))
mclZn <- multcompLetters(fullPTable(pwZn$p.value)) 
# plot compact letters at mean values
pu <- par("usr")
text(c(1:4), rep(pu[3]+0.06*(pu[4]-pu[3]),4), 
           labels=mclZn$Letters, family="serif", font=3)

```

&nbsp;

Figure 3 shows two options for presenting compact letters on a boxplot. You can
choose the one you like best, or adapt the code to make your own!

&nbsp;

## References and R Packages

Garnier S, Ross N, Rudis R, Camargo AP, Sciaini M, Scherer C (2021). *Rvision - Colorblind-Friendly Color Maps for R*. R package version 0.6.2. [https://sjmgarnier.github.io/viridis/](https://sjmgarnier.github.io/viridis/){target="_blank"} (`viridis` package).

Graves S, Piepho H, Dorai-Raj LSwhfS (2024). `multcompView`: *Visualizations of Paired Comparisons*. R package version 0.1-10, [https://CRAN.R-project.org/package=multcompView](https://CRAN.R-project.org/package=multcompView){target="_blank"}.

Mangiafico, Salvatore S. (2024). `rcompanion`: *Functions to Support Extension Education Program Evaluation*. version 2.4.36. Rutgers Cooperative Extension. New Brunswick, New Jersey. [https://CRAN.R-project.org/package=rcompanion](https://CRAN.R-project.org/package=rcompanion){target="_blank"}

Rate, A. W., & McGrath, G. S. (2022). Data for assessment of sediment, soil, and water quality at Ashfield Flats Reserve, Western Australia. *Data in Brief*, **41**, 107970. [https://doi.org/10.1016/j.dib.2022.107970](https://doi.org/10.1016/j.dib.2022.107970){target="_blank"} 

Snow G (2024). `TeachingDemos`: *Demonstrations for Teaching and Learning*. R package version 2.13, [https://CRAN.R-project.org/package=TeachingDemos](https://CRAN.R-project.org/package=TeachingDemos){target="_blank"}
  
Water Quality Australia. (2024). *Toxicant default guideline values for sediment quality.* Department of Climate Change, Energy, the Environment and Water, Government of Australia. Retrieved 2024-04-11 from [https://www.waterquality.gov.au/anz-guidelines/guideline-values/default/sediment-quality-toxicants](https://www.waterquality.gov.au/anz-guidelines/guideline-values/default/sediment-quality-toxicants){target="_blank"}

&nbsp;
