---
title: "R markdown chunk options"
subtitle: "More control over reports"
author: "Andrew Rate"
date: "`r Sys.Date()`"
code_download: TRUE
output: 
  html_document:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
library(flextable) 
library(officer)
set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 10, fonts_ignore = TRUE)
BorderDk <- fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- fp_border(color = "#FFFFFF", style = "solid", width = 1)
```

This brief guide follows on from a previous page on this site, the [Brief Essentials of R Markdown](One-page_R-Markdown.html)

## Hiding both code and output

If we want R code to run, but not show either the code nor any output, we 
include the option `include=FALSE` 

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;">
<span style="color: #644A9B;">&grave;&grave;&grave;{r hide-code-and-output, <span style="background-color: #FF8;">include=FALSE</span>}</span><br>
*&num; the R markdown chunk option include = FALSE is used when it's not*<br>
*&num; helpful to see the code and output, for example loading packages*<br>
<span style="color: #644A9B;">library</span>(car)<br>
<span style="color: #644A9B;">library</span>(lmtest)<br>
<span style="color: #644A9B;">library</span>(robustbase)<br>
<span style="color: #644A9B;">&grave;&grave;&grave;</span></div>

```{r hide-code-and-output, warning=FALSE, message=FALSE, include=FALSE}
# the R markdown chunk option include = FALSE is used when it's not
# helpful to see the code and output, for example loading packages
library(car)
library(lmtest)
library(robustbase)
```

&nbsp;

**Chunk Names**: You'll notice that after `{r ` we include a name like 
`hide-code-and-output`. These names must be unique for each chunk, preferably 
with words separated with hyphens `-`.

Usually we don't want warnings or messages from our `R` code in reports either, 
so we would also include `warning=FALSE` and `include=FALSE`:

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;">
&grave;&grave;&grave;{r hide-code-output-warnings-messages, include=FALSE, <span style="background-color: #FF8;">warning=FALSE, message=FALSE</span>}<br>
⋮<br>
&grave;&grave;&grave;</div>

&nbsp;

## Hiding the code

To hide just the code, but display the output (text-based output, plots, *etc*.),
we include the option `echo=FALSE`. In reports, we normally don't want to 
display the code or any warnings or messages, so we could include all these 
options in each chunk:

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;">
<span style="color: #644A9B;">&grave;&grave;&grave;{r co2-plot-hide-code, <span style="background-color: #FF8;">echo=FALSE</span>, warning=FALSE, message=FALSE}</span><br>
<span style="color: #644A9B;">par</span>(mar=c(<span style="color: #B08000;">3,3,1,1</span>), mgp=c(<span style="color: #B08000;">1.7,0.3,0</span>), tcl=0.2)<br>
<span style="color: #888888;"># co2 is a built-in R dataset</span><br>
<span style="color: #644A9B;">plot</span>(co2, <span style="color: #0057AE;">col</span>="<span style="color: silver; background-color: #0000CD;">blue3</span>", <span style="color: #0057AE;">type</span>=<span style="color: #BF0303;">"l"</span>, <span style="color: #0057AE;">xlab</span>=<span style="color: #BF0303;">"Date"</span>,<br>
&nbsp; &nbsp; &nbsp;<span style="color: #0057AE;">ylab</span>=<span style="color: #644A9B;">expression</span>(<span style="color: #644A9B;">paste</span>(<span style="color: #BF0303;">"Atmospheric "</span>,CO[<span style="color: blue;">2</span>],<span style="color: #BF0303;">" concentration at Mauna Loa"</span>)))<br>
<span style="color: #644A9B;">&grave;&grave;&grave;</span></div>

&nbsp; 

```{r co2-plot-hide-code, echo=FALSE}
par(mar=c(3,3,1,1), mgp=c(1.7,0.3,0), tcl=0.2)
plot(co2, col="blue3", type="l", xlab="Date", 
     ylab=expression(paste("Atmospheric ",CO[2]," concentration (ppm) at Mauna Loa")))
```

&nbsp;

A more efficient way to do this, however, would be to modify the chunk at the 
beginning of our markdown when we create a new R markdown file in R:
<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;">
<span style="color: #0057AE;">&grave;&grave;&grave;{r set-knitr-chunk-options}</span><br>
knitr::<span style="color: #644A9B;">opts_chunk</span><span style="color: #3DAEE9;">$</span>set(<span style="color: #0057AE;">echo</span>=<span style="color: #AA5500;">FALSE</span>, <span style="color: #0057AE;">warning</span>=<span style="color: #AA5500;">FALSE</span>, <span style="color: #0057AE;">message</span>=<span style="color: #AA5500;">FALSE</span>)<br>
<span style="color: #644A9B;">&grave;&grave;&grave;</span></div>

&nbsp;

If we set these chunk options at the beginning, we can omit the options 
`echo=FALSE, warning=FALSE, message=FALSE` in our individual code chunks. 
We can override them if we want, *e.g*. by including `echo=TRUE`.

## Re-sizing plots

One of the most useful sets of chunk options allows to change how plots appear,
such as `fig.width`, `fig.height`, `fig.align`, and so on. For example, we could
use `fig.width` and `fig.height` make the previous plot squarer:

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;">
<span style="color: #644A9B;">&grave;&grave;&grave;{r co2-plot-squarer, echo=FALSE, warning=FALSE, message=FALSE, <span style="background-color: #FF8;">fig.width=5, fig.height=5</span>}</span><br>
⋮<br>
<span style="color: #644A9B;">&grave;&grave;&grave;</span></div>

```{r co2-plot-squarer, echo=FALSE, fig.width=5, fig.height=5}
par(mar=c(3,3,1,1), mgp=c(1.7,0.3,0), tcl=0.2)
plot(co2, col="blue3", type="l", xlab="Date", 
     ylab=expression(paste("Atmospheric ",CO[2]," concentration (ppm) at Mauna Loa")))
```

&nbsp; 

Setting `fig.width` and `fig.height` is for the whole plot area so, if we have a 
multiple-frame plot (*i.e*. using `mfrow=` or `mfcol=` in the `par()` function), we would need to adjust the dimensions to match. For example:

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;"><span style="color: #644a9b;">&grave;&grave;&grave;{r co2-and sunspot-plots, echo=FALSE, warning=FALSE, message=FALSE, <span style="background-color: #ff8;">fig.width=10, fig.height=5, fig.cap="Plots of two time series: atmospheric CO₂ concentrations (left), and sunspot numbers (right)."</span>}</span><br /><span style="color: #644a9b;">par</span>(<span style="color: #0057ae;">mfrow</span>=c(<span style="color: #b08000;">1,2</span>), <span style="color: #0057ae;">mar</span>=c(<span style="color: #b08000;">3,3,1,1</span>), <span style="color: #0057ae;">mgp</span>=c(<span style="color: #b08000;">1.7,0.3,0</span>), <span style="color: #0057ae;">tcl</span>=<span style="color: #b08000;">0.2</span>, <span style="color: #0057ae;">font.lab</span>=<span style="color: #b08000;">2</span>)<br /><span style="color: #888888;"># co2 and sunspots are built-in R datasets</span><br /><span style="color: #644a9b;">plot</span>(co2, <span style="color: #0057ae;">col</span>="<span style="color: silver; background-color: #0000cd;">blue3</span>", <span style="color: #0057ae;">type</span>=<span style="color: #bf0303;">"l"</span>, <span style="color: #0057ae;">xlab</span>=<span style="color: #bf0303;">"Date"</span>,<br />&nbsp; &nbsp; &nbsp;<span style="color: #0057ae;">ylab</span>=<span style="color: #644a9b;">expression</span>(<span style="color: #644a9b;">bold</span>(<span style="color: #644a9b;">paste</span>(<span style="color: #bf0303;">"Atmospheric "</span>,CO[<span style="color: blue;">2</span>],<span style="color: #bf0303;">" concentration at Mauna Loa"</span>))),&nbsp;<br />&nbsp; &nbsp; &nbsp;<span style="color: #0057ae;">main</span>=<span style="color: #644a9b;">expression</span>(<span style="color: #644a9b;">paste</span>(<span style="color: #bf0303;">"Atmospheric "</span>,CO[<span style="color: blue;">2</span>],<span style="color: #bf0303;">" concentration at Mauna Loa"</span>)))<br /><span style="color: #644a9b;">plot</span>(sunspots, <span style="color: #0057ae;">col</span>="<span style="background-color: #b22222; color: #d3d3d3;">firebrick</span>", <span style="color: #0057ae;">type</span>=<span style="color: #bf0303;">"l"</span>, <span style="color: #0057ae;">xlab</span>=<span style="color: #bf0303;">"Date"</span>,<br />&nbsp; &nbsp; &nbsp;<span style="color: #0057ae;">ylab</span>=<span style="color: #bf0303;">"Relative number of sunspots"</span>, <span style="color: #0057ae;">main</span>=<span style="color: #bf0303;">"Monthly Sunspot Numbers, 1749&ndash;1983"</span>)<br /><span style="color: #644a9b;">&grave;&grave;&grave;</span></div>

&nbsp; 

```{r co2-and sunspot-plots, echo=FALSE, fig.width=12, fig.height=5, fig.cap="Plots of two time series: atmospheric CO₂ concentrations (left), and sunspot numbers (right)."}
par(mfrow=c(1,2), mar=c(3,3,2,1), mgp=c(1.7,0.3,0), tcl=0.2, font.lab=2)
plot(co2, col="blue3", type="l", xlab="Date", 
     ylab=expression(bold(paste(CO[2]," (ppm)"))), 
     main=expression(paste("Atmospheric ",CO[2]," concentration at Mauna Loa")))
plot(sunspots, col="firebrick", type="l", xlab="Date",
     ylab="Relative number of sunspots", main="Monthly Sunspot Numbers, 1749–1983")
```

&nbsp;

## Adding figure captions to plots

We add captions using `fig.caption="<figure caption text>"`. This is a better 
option for reporting than including a title above plots using the `main=` option

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;">
<span style="color: #644A9B;">&grave;&grave;&grave;{r co2-plot-caption, echo=FALSE, warning=FALSE, message=FALSE, fig.width=5, fig.height=5, <span style="background-color: #FF8;">fig.align='center', fig.cap="Figure 1: Time series of atmospheric CO₂ concentrations from 1959 to 1998, measured at the Mauna Loa observatory in Hawaii."</span>}</span><br>
⋮<br>
<span style="color: #644A9B;">&grave;&grave;&grave;</span></div>

```{r co2-plot-caption, echo=FALSE, fig.width=5, fig.height=5, fig.align='center', fig.cap="Figure 1: Time series of atmospheric CO₂ concentrations from 1959 to 1998, measured at the Mauna Loa observatory in Hawaii."}
par(mar=c(3,3,1,1), mgp=c(1.7,0.3,0), tcl=0.2)
plot(co2, col="blue3", type="l", xlab="Date", 
     ylab=expression(paste("Atmospheric ",CO[2]," concentration (ppm) at Mauna Loa")))
```

&nbsp; 

We could just use normal markdown text to insert a caption, but using 
`fig.caption` can allow automatic cross-referencing of figures and tables to 
their captions using the `bookdown` R package.

At this stage it's useful to note that with a long chunk header (within `{}`),
such as that shown above, we should not insert any line breaks. The chunk will
return an error and not work unless there are *no line breaks* in the chunk
header.

&nbsp;

## Using the `bookdown` package

The R package `bookdown` is very powerful, but so far I only use it for its 
ability to provide automatic Figure and Table numbering, and cross-referencing.

You would need to run the following R code if you don't have `bookdown` installed:

```{r install-bookdown, eval=FALSE, echo=TRUE}
if(!require(bookdown)) install.packages("bookdown")
```

You would also need to include the following in the `yaml` header at the top of your R markdown document:

```
---
⋮
output: 
  bookdown::html_document2:
⋮
---
```

This should give a menu option to `Knit to html_document2`, which is what we
should use later to prepare our formatted document. 

We can then use chunk names to cross reference figures and tables. For
example, for the chunk above plotting two figures...

<div style="border: 2px solid #cccccc; padding: 5px; font-family: 'Courier New', monospace;"><span style="color: #644a9b;">&grave;&grave;&grave;{r co2-and sunspot-plots, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=5, <span style="background-color: #ff8;">fig.cap="Plots of two time series: atmospheric CO₂ concentrations (left), and sunspot numbers (right)."</span>}</span><br />
⋮<br>
<span style="color: #644A9B;">&grave;&grave;&grave;</span></div>

<br>...we would cross-reference this by including `Figure \@ref(fig:co2-and sunspot-plots)` at the appropriate place in our text.

**Notes** 

1. chunk names can not contain spaces, periods, underscores, or slashes; words should just be separated by hyphens (`-`), and all chunk names must be unique
2. we don't need to include the word `Figure` or a figure number at the start of 
the `fig.cap=` text string in the chunk header (`{r . . . }`)

Similarly, table captions can be cross-referenced using 
`Table \@ref(tab:table-chunk-name)` in the text outside of any code chunks. We
would need to set a table caption in the chunk header using `tab.cap=`, or use a
package like `flextable` or `knitr::kable()` and include table caption options
from those packages.

&nbsp;

## Other chunk options

```{r other-chunk-options, message=FALSE, warning=FALSE, paged.print=FALSE, echo=FALSE}
optsTable <- data.frame(Option=c("out.width=",
                                 "fig.align=",
                                 "results='hold'",
                                 "paged.print="),
                        Purpose=c("Sets the width of a figure as a proportion of the output page width,\ne.g. out.width='50%'",
                                  "Aligns the figure on the output page\ne.g. fig.align='center'",
                                  "Waits until all the code in a chunk has run before printing any text output\n(other choices are 'markup', 'hide', 'asis')",
                                  "If TRUE (the default) adds html format to printed data frames;\nif FALSE output is just text"))
flextable(optsTable) |> 
  width(width=c(2,6)) |> valign(valign="top", part="body") |> 
  font(j=1, fontname = 'Courier New') |> 
  color(j=1, color="#644A9B") |> 
  border_outer(border = BorderDk, part = "all") |>
  border_inner_v(border=BorderDk, part="header") |> 
  set_caption(caption="Table 1: Some useful options for controlling the output from code chunks in R markdown.", align_with_table=F, fp_p=fp_par(text.align = "left", padding.bottom = 6))
```

You can see a more complete table in the [R Markdown Cheatsheet](https://rstudio.github.io/cheatsheets/html/rmarkdown.html#chunk-options){target="_blank"}.

&nbsp;

## Other R Markdown Resources

* [R Markdown Cheat Sheet](https://rstudio.github.io/cheatsheets/html/rmarkdown.html){target="_blank"} (Posit Software, 2024)

* [Getting used to R, RStudio, and R Markdown](https://bookdown.org/chesterismay/rbasics/4-rmarkdown.html){target="_blank"} – Chapter 4 in the excellent (and free) eBook by [Chester Ismay](https://github.com/ismayc){target="_blank"}

* [Chunk Options and Package Options from Yihui Xie's website](https://yihui.org/knitr/options/){target="_blank"}

* [Chunk Options – Chapter 11 in R markdown cookbook by Xie *et al*. (2025)](https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html){target="_blank"}

### References

Ismay, Chester (2016) *Getting used to R, RStudio, and R Markdown*. [https://bookdown.org/chesterismay/rbasics/](https://bookdown.org/chesterismay/rbasics/){target="_blank"} (accessed 2025-06-17)

Posit Software (2024) `rmarkdown::`CHEATSHEET. [https://rstudio.github.io/cheatsheets/rmarkdown.pdf](https://rstudio.github.io/cheatsheets/rmarkdown.pdf){target="_blank"} (accessed 2025-06-17)

Xie, Yihui (2020) *Chunk Options and Package Options*. [https://yihui.org/knitr/options/](https://yihui.org/knitr/options/){target="_blank"} (accessed 2025-06-17)

Xie Y, Dervieux C, Riederer E (2025) *R Markdown Cookbook*. [https://bookdown.org/yihui/rmarkdown-cookbook/](https://bookdown.org/yihui/rmarkdown-cookbook/){target="_blank"} (accessed 2025-06-17)
