---
title: "Using R for soil profile diagrams"
subtitle: "A selection of methods"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document: 
    highlight: kate
---

We're going to illustrate all the methods using some made-up data. You would need to adapt the **R** code examples to match your dataset and variable names.

<div style="border: 2px solid #039; background-color:#e8e8e8; padding: 8px;">
For more options for plotting depth profiles, including for multiple cores, go to [**this page**](depth-profiles-more.html).
</div>

&nbsp;

Just so we don't need to read in a file, we're going to make a dataframe using R code:

```{r make the data, results='hold'}
profile <- data.frame(Sample = seq(1,10),
                      Horizon = c("A","A","A2","A2","B1","B1","B1","B2","B2","B3"),
                      Depth_upper = c(0,5,9,14,20,26,31,35,43,49),
                      Depth_lower = c(5,9,14,20,26,31,35,43,49,55),
                      Sand = c(81.8, 89.6, 80.4, 72.9, 45.1, 65.5, 56.6, 55.9, 39.4, 37.9),
                      Silt = c(13.8, 4.3, 5.8, 6.0, 18.2, 13.8, 15.8, 14.0, 21.7, 19.5),
                      Clay = c(4.4, 6.1, 13.8, 21.1, 36.7, 20.7, 27.6, 30.1, 38.9, 42.6))
profile$Horizon <- as.factor(profile$Horizon)
str(profile)
cat("\nSum of sand + silt + clay in each sample:\n")
with(profile, Sand+Silt+Clay)
```

## Inverted scatter plot (the most basic option)

To invert a scatter plot t make it suitable for soil depth profile plotting, we:

-   use [par(mar = c( . . . ))]{style="font-family: monospace; background-color:#e8e8e8;"} to make the upper margin larger to fit in axis and tick labels
-   calculate mean depth for each sample = (upper + lower)/2
-   plot Depth *vs*. variable (which is inverted)
-   suppress the x axis ([xaxt="n"]{style="font-family: monospace; background-color:#e8e8e8;"}) and have no x-axis label ([xlab=""]{style="font-family: monospace; background-color:#e8e8e8;"})
-   plot the depth axis backwards by putting the deepest limit first ([ylim=c(55,0)]{style="font-family: monospace; background-color:#e8e8e8;"})
-   rotate the labels using the [las = 1]{style="font-family: monospace; background-color:#e8e8e8;"} option in [par()]{style="font-family: monospace; background-color:#e8e8e8;"} or [plot()]{style="font-family: monospace; background-color:#e8e8e8;"}
-   add a manual x axis at the top using [axis(3)]{style="font-family: monospace; background-color:#e8e8e8;"}
-   add a manual x axis title at top using the [mtext()]{style="font-family: monospace; background-color:#e8e8e8;"} function

```{r set default color palette invisibly, message=FALSE, warning=FALSE, include=FALSE}
palette("default")
```

```{r simple depth profile, fig.height=5, fig.width=3, echo=1:7, results='hold', fig.align='center', fig.cap="Figure 1. Depth profile of clay in soil as a scatter plot with inverted axes"}
# omit axis plotting and axis label...
# ... then plot axis manually, first adjusting plot margins
par(mar=c(1,3,3,1), mgp = c(1.6,0.3,0), tcl = 0.3, font.lab = 2) # won't change anything except margins
plot(profile$Clay, ((profile$Depth_upper+profile$Depth_lower)/2),
     xlab="", xaxt="n", type = "b", las = 1,
     xlim = c(0,90), ylim = c(55,0), 
     ylab="Mean depth (cm)")
# we need two functions / code lines for manual axis, plus label
axis(3)
mtext(side=3, line=1.6, font=2, text="Clay (% of <2mm fraction)")
rect(45,18,80,9)
text(c(62.5,45,62.5,80,62.5), c(18,13.5,9,13.5,9), 
     labels = c("1","2","3","4","\nAxis\nnumbers"), 
     pos = c(1,2,3,4,1), font=c(1,1,1,1,2),col=c(1,1,1,1,4), offset = 0.2)
```

The depth profile plot in Figure 1 is quite a common way to represent soil profile data. It has the advantage of being relatively simple, but implies a continuous change in the variable (Clay in Figure 1) which we usually do not have data for, since we collect samples in discrete depth increments. We will address how to deal with discrete increments later.

Unless we have replicate samples from the same depth increment, we don't have an estimate of sampling+measurement uncertainty. The next section uses an interpolation method to generate a standard error around a smoothed depth profile relationship.

## Smoothed Depth Profiles with uncertainties

To generate smoothed curves of soil properties down a profile, we:

-   first plot just the measurements *vs*. mean depth as above
-   create an 'empty' data frame to hold the smoothed interpolated results
-   for each series, create a smoothing model using the [loess()]{style="font-family: monospace; background-color:#e8e8e8;"} function (*LOESS* is <u>L</u>ocally <u>E</u>stimated <u>S</u>catterplot <u>S</u>moothing)
-   generate an object containing predictions and standard errors from the LOESS models using the [predict()]{style="font-family: monospace; background-color:#e8e8e8;"} function
-   plot the smooth (interpolated) lines and their standard errors
-   add a legend

```{r interp depth profile, fig.height=5, fig.width=3, results='hold', fig.align='center', fig.cap="Figure 2. Depth profiles of clay and sand in soil as smoothed scatter plots with standard error intervals; actual measurements are shown by plot symbols."}
par(mar=c(1,3,3,1), mgp = c(1.6,0.3,0), tcl = 0.3, font.lab = 2, lend=2, ljoin=1)
profile$Depth_mean <- (profile$Depth_upper + profile$Depth_lower)/2
with(profile, plot(Depth_mean ~ Clay, xaxt = "n", xlab = "", ylab = "Mean depth (cm)",
                   pch = 19, col = 4, ylim = c(55,0), xlim = c(0,90), 
                   cex = 1.2, cex.lab = 1.5))
with(profile, points(Depth_mean ~ Sand, pch = 0, col = 2, lwd=2))
grid()
axis(3)
mtext(expression(bold(paste("Size fraction (%)"))),
      side = 3, line = 1.5, font = 2)

# set up data.frame to store interpolation results
interps <- data.frame(fixdepth = seq(2.5,60,5),
                      Clay = rep(NA, length(seq(2.5,60,5))),
                      Clay_lo = rep(NA, length(seq(2.5,60,5))),
                      Clay_hi = rep(NA, length(seq(2.5,60,5))),
                      Sand = rep(NA, length(seq(2.5,60,5))),
                      Sand_lo = rep(NA, length(seq(2.5,60,5))),
                      Sand_hi = rep(NA, length(seq(2.5,60,5))))

# generate Clay values interpolated to fixed depth using LOESS
loessMod1 <- loess(Clay ~ Depth_mean, data = profile)
loessPred1 <- predict(loessMod1, newdata=interps$fixdepth, se = TRUE)
interps$Clay <- loessPred1$fit
interps$Clay_lo <- loessPred1$fit - loessPred1$se.fit
interps$Clay_hi <- loessPred1$fit + loessPred1$se.fit

# generate Core 2 values interpolated to fixed depth using LOESS
loessMod2 <- loess(Sand ~ Depth_mean, profile)
loessPred2 <- predict(loessMod2, newdata=interps$fixdepth, se = TRUE)
interps$Sand <- loessPred2$fit
interps$Sand_lo <- loessPred2$fit - loessPred2$se.fit
interps$Sand_hi <- loessPred2$fit + loessPred2$se.fit

# add interpolations to plot
lines(interps$Clay, interps$fixdepth, col = 4)
lines(interps$Clay_lo, interps$fixdepth, col = 4, lty = 2, lwd = 2)
lines(interps$Clay_hi, interps$fixdepth, col = 4, lty = 2, lwd = 2)

lines(interps$Sand, interps$fixdepth, col = 2)
lines(interps$Sand_lo, interps$fixdepth, col = 2, lty = 2, lwd = 2)
lines(interps$Sand_hi, interps$fixdepth, col = 2, lty = 2, lwd = 2)

legend("top", bty = "o", inset = 0.02,
       cex = 0.8, box.col = "gray", box.lwd = 1, 
       legend = c("Clay", "", "Sand",""), 
       col = c(4,4,2,2), lty = c(1,2,1,2), lwd = c(1,2,1,2), pt.lwd = 2, 
       pch = c(19,NA,0,NA), pt.cex = c(1.2,NA,1,NA))
```

The depth profile plot in Figure 2 is a nice way to represent soil profile data. It tries to predict the uncertainty in measurement with depth, assuming a continuous smooth relationship (which might not be a valid assumption!). We can plot our depth profile with no assumptions about how out variable(s) change between measurements using a variation on a bar plot.

## Soil depth profile plots as cumulative stepped plots

To make stepped plots makes use of the [type = "s"]{style="font-family: monospace; background-color:#e8e8e8;"} option in the functions [plot()]{style="font-family: monospace; background-color:#e8e8e8;"} or [lines()]{style="font-family: monospace; background-color:#e8e8e8;"}:

```{r stepped depth profile, fig.height=5, fig.width=3, results='hold', fig.align='center', fig.cap="Figure 3. Depth profile of clay, silt, and sand in soil as a stepped line plot."}
par(mar=c(1,3,3,1), mgp = c(1.6,0.3,0), tcl = 0.3, font.lab = 2, lend=2, ljoin=1, las = 1)
plot(c(0, profile$Depth_mean) ~ c(profile$Clay[1], profile$Clay), 
     type = "s", xlim = c(0,105), ylim = c(70,0),
     pch = 16, xaxt = "n", ylab = "Depth (cm)")
lines(c(0,profile$Depth_mean) ~ c(c(profile$Silt[1],profile$Silt) +
                                    c(profile$Clay[1], profile$Clay)), 
      type="s", col="red3", lty = 2)
lines(c(0,profile$Depth_mean) ~ c(c(profile$Sand[1],profile$Sand) +
                                    c(profile$Silt[1],profile$Silt) +
                                    c(profile$Clay[1], profile$Clay)), 
      type="s", col="blue3", lty = 3, lwd = 2, lend="round")
axis(3); mtext("Cumulative (%)", side=3, line=1.6, font=2, cex=1)
legend("bottom", legend = c("Clay","Clay+Silt","Clay+Silt+Sand"),
       lty = c(1,2,3), lwd = c(1,1,2), col = c("black","red3","blue3"), bty = "n")
```

The plot in Figure 3 is a more "honest" representation of the way that we typically present soil profile data, but it is somewhat hard to interpret as drawn (a cumulative composition plot), and the legend is messy. It would be better to represent the stepped lines as polygons, similar to stacked bars.

## Soil depth profile plots as **filled** cumulative stacked plots

This type of plot obviously works best for fixed-sum compositions, where we have measured all the components, such as soil texture (sand, silt, clay).

The basis of this plot is to use loops to draw rectangles for each series using the [rect()]{style="font-family: monospace; background-color:#e8e8e8;"} function. Except for the first series, we add the previous series so the rectangles 'stack' horizontally. We add stepped lines with [type = "s"]{style="font-family: monospace; background-color:#e8e8e8;"} to outline the adjoined boxes for each series. We use a [layout]{style="font-family: monospace; background-color:#e8e8e8;"} so we can plot in frames of different sizes (in this example, **2** frames, with the left twice as wide as the right). In the right frame we plot a dummy plot with no plot elements drawn, and over this we draw a nice legend (it's probably also possible with a wide right margin, if [xpd = TRUE]{style="font-family: monospace; background-color:#e8e8e8;"} is set in [par()]{style="font-family: monospace; background-color:#e8e8e8;"}).

```{r stacked depth profile, fig.height=5, fig.width=4.5, results='hold', fig.align='center', fig.cap="Figure 4. Depth profiles of clay, silt, and sand in soil as a horizontal stacked plot."}
par(mar=c(1,3.5,3.5,1), mgp = c(1.7,0.3,0), tcl = 0.25, font.lab = 2, 
    las = 1, xaxs="i", yaxs="i", xpd = TRUE)
palette(c("black", 
          colorRampPalette(c("antiquewhite","chocolate4"), space = "rgb")(3),"sienna", 
          colorRampPalette(c("blue3","white"), space = "rgb")(4),"transparent"))
layout(matrix(c(1,1,2), ncol = 3))
plot(c(0, profile$Depth_lower) ~ c(profile$Clay[1], profile$Clay), 
     type = "n", xlim = c(0,100), ylim = c(55,0), ylab = "Depth (cm)",
     cex.lab = 1.5, cex.axis = 1.2, pch = 16, xaxt = "n")
axis(3, cex.axis = 1.2, mgp=c(1.7,0.3,0))
mtext("Fraction of fine earth (%)", side=3, line=1.6, family="sans", font=2, cex=1)

for (i in 1:NROW(profile)) {
  rect(0, profile[i, "Depth_lower"], 
       profile[i,"Clay"], profile[i, "Depth_upper"],
       col = 2, border = 10)}

for (i in 1:NROW(profile)) {
  rect(profile[i,'Clay'], profile[i, "Depth_lower"], 
       profile[i,"Clay"]+profile[i,"Silt"], profile[i, "Depth_upper"],
       col = 3, border = 10)}

for (i in 1:NROW(profile)) {
  rect(profile[i,'Clay']+profile[i,"Silt"], profile[i, "Depth_lower"], 
       profile[i,"Clay"]+profile[i,"Silt"]+profile[i,"Sand"], profile[i, "Depth_upper"],
       col = 4, border = 10)}

box()
axis(2, at = seq(10,50,10), labels = rep(NA, length(seq(10,50,10))), tcl=0.4)
axis(3, at = seq(20,80,20), labels = rep(NA, length(seq(20,80,20))), tcl=0.4)
lines(c(0, profile$Depth_lower) ~ c(profile$Clay[1], profile$Clay), 
      type = "s")
lines(c(0,profile$Depth_lower) ~ c(c(profile$Silt[1],profile$Silt) +
                                  c(profile$Clay[1], profile$Clay)), 
      type="s", lty = 1)
lines(c(0,profile$Depth_lower) ~ c(c(profile$Sand[1],profile$Sand) +
                                  c(profile$Silt[1],profile$Silt) +
                                  c(profile$Clay[1], profile$Clay)), 
      type="s", lty = 1)

plot(c(0, 1), c(0, 1), axes=F, ann=F, type="n")
legend(0.05, 0.65, legend = c("Clay","Silt","Sand"), 
       cex = 2, pch = 22, pt.bg=c(2,3,4), pt.cex = 8,
       bty="n", bg = 9, box.col = 1, inset=0,
       xjust=0.5, x.intersp=1.2, y.intersp=1.5)
```

The resulting horizontally stacked plot in Figure 4 also presents a more "honest" representation of soil profile data, and it is easier to interpret than Figure 3, with a much tidier legend. It would be hard to indicate measurement variability on this type of plot, but it's potentially an attractive option.

```{r art, echo=FALSE, fig.height=5, fig.width=7.5, results='hold', fig.align='center', fig.cap="Some soil-themed geneRative R art (Rtistry) just for fun and decoration."}
par(mar = c(.1,.1,.1,.1), xpd=T)
plot(c(0,1),c(0,1), axes=F, type="n",ann=F)
for (i in 1:100) {
  symbols(sample(seq(0,1,l=100),100), jitter(seq(0,1,l = 100)), circles=sample(seq(0,.01,l=100),100), add=T,
          fg="transparent", 
          bg=colorRampPalette(c("antiquewhite","chocolate4"), space = "rgb")(33), lwd=2, inches=F)
}
```
