---
title: "Using R for soil profile diagrams"
subtitle: "Additional methods"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document: 
    highlight: kate
---

```{r setup-and-packages, include=FALSE, message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
library(car)
library(viridis)
library(TeachingDemos)
library(rcompanion)
library(multcompView)
library(ggplot2)})
```

We're going to illustrate some more methods for drawing depth profiles and their
trends using some real sediment core data. As [before](depth-profiles.html), you
would need to adapt the **R** code examples to match your dataset and variable
names.

First we read in a file and do some corrections of the data:

```{r}
git <- 
  "https://raw.githubusercontent.com/Ratey-AtUWA/Learn-R-web/refs/heads/main/"
afs25 <- read.csv(paste0(git,"afs25.csv"), stringsAsFactors = TRUE)
afs25$Unit <- as.factor(afs25$Unit)
afs25$Group <- as.factor(afs25$Group)
afs25$SampleID <- as.character(afs25$SampleID)
afs25$Notes <- as.character(afs25$Notes)
```

## Boxplots in depth categories

This may be a useful approach if we want to summarise depth trends across a site
or stratum, especially if profiles are of different depths (*i.e*.
under-representation of greater depths).

We need the `rcompanion` package for the `fullPTable()` function, the
`multcompView` package for the `multcompLetters()` function, and the
`TeachingDemos` package for the `shadowtext()` function.

```{r depth-factor, fig.height=6, fig.width=6, message=FALSE, warning=FALSE, fig.cap="Figure 1: Depth profile of copper in multiple sediment cores, shown as boxplots in depth categories. Mean values are shown as blue '+' symbols; the letters above symbols differ if pairwise Wilcoxon test comparisons are significant (p ≤ 0.05). Bold numbers under '+' symbols are mean values."}
afs25$DepthF <- cut(afs25$Depth_mean, breaks=c(0,10,20,30,100), 
                    labels = c("0-10cm","10-20cm","20-30cm",">30cm"))
par(mfrow=c(1,1), mar=c(1,5,3,1), mgp=c(1.6,0.4,0), tcl=-0.2, lend="square",
    font.lab=2)
palette(c("black","blue3","navy","red3","#ffa00040","white","transparent"))
with(afs25, boxplot(log10(Cu) ~ DepthF, horizontal=T, boxwex=1, ylab="", 
                    xaxt="n", xlab="", xaxs="i", col=5, 
                    cex=1.2, xlim=c(4.3,0.7), las=1, xaxt="n"))
abline(h=seq(1.5,3.5), lty=2, col="#00000040")
axis(3, at=log10(c(1,2,5,10,20,50,100,200,500)), 
          labels=c(1,2,5,10,20,50,100,200,500))
mtext("Cu (mg/kg)", side=3, line=1.6, font=2)
mtext("Depth Range", side=2, line=4, font=2)
meanz <- with(afs25, tapply(log10(Cu), DepthF, mean, na.rm=T))
points(meanz, 1:nlevels(afs25$DepthF), pch=3, col=6, lwd=4)
points(meanz, 1:nlevels(afs25$DepthF), pch=3, col=2, lwd=2)
kwt <- kruskal.test(Cu~DepthF, data=afs25)
pwt <- with(afs25, pairwise.wilcox.test(Cu, DepthF))
fpt <- fullPTable(pwt$p.value)
cld <- multcompLetters(fpt)
shadowtext(meanz, 1:nlevels(afs25$DepthF), labels=signif(meanz,3), 
           font=2, family="sans", pos=1, col=3, bg=6,r=0.2)
shadowtext(meanz, 1:nlevels(afs25$DepthF), labels=cld[[2]], 
           font=3, family="serif", pos=3, col=2, bg=6, r=0.2)
text(max(log10(afs25$Cu), na.rm=T), 4.3,
     pos=2, col=3, labels=paste("Kruskal-Wallis p =",signif(kwt$p.value,2)))
```

&nbsp;

## Variable as a linear function of depth – many profiles

This may be another useful way to summarise depth trends, again especially if
profiles are of different depths (*i.e*. not all depths are equally
represented).

We do everything in base-R without additional packages.

```{r depth-profile-lm, fig.height=6, fig.width=4, message=FALSE, warning=FALSE, fig.cap="Figure 2: Depth profile of copper concentrations for multiple sediment cores, with Cu concentrations fitted to a linear model with Depth as the single predictor."}
par(mfrow=c(1,1), mar=c(1,3,3,1), mgp=c(1.6,0.4,0), tcl=-0.2, lend="square",
    font.lab=2)
palette(c("black","blue3","red3","#ff000040","#ffffffc0","transparent"))
with(afs25, plot(log10(Cu), Depth_mean, type="n", ylab="",
                 xaxt="n", xlab="", yaxs="i", cex=1.2, 
                 xlim=log10(c(3,640)), ylim=c(70,0), 
                 las=1, xaxt="n"))
abline(h=seq(10,60,10), lty=2, col="#00000040")
axis(3, at=log10(c(1,2,5,10,20,50,100,200,500)), 
          labels=c(1,2,5,10,20,50,100,200,500))
mtext("Cu (mg/kg)", side=3, line=1.6, font=2)
mtext("Depth (cm)", side=2, line=1.6, font=2)
lm0 <- lm(log10(Cu)~Depth_mean, data=afs25)
slm0 <- summary(lm0)
depfz <- data.frame(Depth_mean = seq(2.5,70,2.5))
predz <- as.data.frame(predict(lm0, newdata = depfz, interval="confidence"))
predz <- cbind(depfz, predz)
polygon(c(predz$lwr, rev(predz$upr), predz$lwr[1]),
        c(predz$Depth_mean, rev(predz$Depth_mean),predz$Depth_mean[1]),
        col=4, border=6)
lines(predz$fit, predz$Depth_mean, col=2)
lines(predz$lwr, predz$Depth_mean, col=3, lty=2)
lines(predz$upr, predz$Depth_mean, col=3, lty=2)
with(afs25, points(log10(Cu), Depth_mean, cex=1.2))
text(par("usr")[2],par("usr")[3],col=2,
     labels=paste0("R² = ",signif(slm0$r.sq,3), "; p = ",
                   signif(1-(pf(summary(lm0)$fst, 1, lm0$df.res)[[1]]),3),
                   "\n"), pos=2)
legend("bottomright", legend=c("Observations","Linear model","Model 95% CI"),
       pch=c(1,NA,15), pt.cex=c(1.2,NA,2), col=c(1,2,4), lwd=c(NA,1,NA),
       bg=5, box.col=6, inset=c(0.005,0.05))
```

&nbsp;

## Variable as a linear function of depth – single profile

We can also apply a linear model to a single depth profile, *e.g*. from one
sediment core or soil pit.

```{r depth-profile-g10-lm, fig.height=6, fig.width=4, message=FALSE, warning=FALSE, fig.cap="Figure 3: Depth profile of gadolinium concentrations for a single sediment core, with Gd concentrations fitted to a linear model with Depth as the single predictor."}
par(mfrow=c(1,1), mar=c(1,3,3,1), mgp=c(1.6,0.4,0), tcl=-0.2, lend="square",
    font.lab=2)
palette(c("black","blue3","red3","#ff000040","#ffffffc0","transparent"))
g10 <- droplevels(afs25[which(afs25$Group==10),])
with(g10, plot(Gd, Depth_mean, type="n", ylab="", 
                    xaxt="n", xlab="", yaxs="i", 
                    cex=1.2, xlim=c(-2,11), ylim=c(70,0), las=1, xaxt="n"))
abline(h=seq(10,60,10), lty=2, col="#00000040")
axis(3)
mtext("Gd (mg/kg)", side=3, line=1.6, font=2)
mtext("Depth (cm)", side=2, line=1.7, font=2)
lm0 <- lm(Gd~Depth_mean, data=g10)
slm0 <- summary(lm0)
depfz <- data.frame(Depth_mean = seq(2.5,70,2.5))
predz <- as.data.frame(predict(lm0, newdata = depfz, interval="confidence"))
predz <- cbind(depfz, predz)
polygon(c(predz$lwr, rev(predz$upr), predz$lwr[1]),
        c(predz$Depth_mean, rev(predz$Depth_mean),predz$Depth_mean[1]),
        col=4, border=6)
lines(predz$fit, predz$Depth_mean, col=2)
lines(predz$lwr, predz$Depth_mean, col=3, lty=2)
lines(predz$upr, predz$Depth_mean, col=3, lty=2)
with(g10, points(Gd, Depth_mean, cex=1.2, pch=19))
text(par("usr")[2],par("usr")[3],col=2,
     labels=paste0("R² = ",signif(slm0$r.sq,2), "; p = ",
                   signif(1-(pf(summary(lm0)$fst, 1, lm0$df.res)[[1]]),2),
                   "\n"), pos=2)
legend("bottomright", legend=c("Observations","Linear model","Model 95% CI"),
       pch=c(19,NA,15), pt.cex=c(1.2,NA,2), col=c(1,2,4), lwd=c(NA,1,NA),
       bg=5, box.col=6, inset=c(0.005,0.05))
```

&nbsp;

## Variable as a cubic polynomial function of depth – single profile

A linear model may not be very satisfying or appropriate for a single profile.
We can adapt a linear model to fit a polynomial curve by creating columns in our
data for `Depth²` and `Depth³`, and expressing our variable as a linear function
of `Depth`, `Depth²`, and `Depth³` using multiple regression. A cubic polynomial
is usually sufficient to capture any curvature.

```{r depth-profile-g10-cubic, fig.height=6, fig.width=4, message=FALSE, warning=FALSE, fig.cap="Figure 4: Depth profile of gadolinium concentrations for a single sediment core, with Gd concentrations fitted to a cubic function of Depth."}
par(mfrow=c(1,1), mar=c(1,3,3,1), mgp=c(1.6,0.4,0), tcl=-0.2, lend="square",
    font.lab=2)
palette(c("black","blue3","red3","#ff000040","#ffffffc0","transparent"))
data0 <- na.omit(afs25[which(afs25$Group==10),c("Depth_mean","Gd")])
data0$Depth2 <- data0$Depth_mean^2
data0$Depth3 <- data0$Depth_mean^3
with(data0, plot(log10(Gd), Depth_mean, col="#00000040", pch=19, ylab="", 
                 xaxt="n", xlab="", yaxs="i", 
                 cex=1.2, xlim=log10(c(1,13)), ylim=c(70,0), las=1, xaxt="n"))
abline(h=seq(10,60,10), lty=2, col="#00000040")
axis(3, at=log10(c(1,2,5,10,20,50,100,200,500)), 
     labels=c(1,2,5,10,20,50,100,200,500))
mtext("Gd (mg/kg)", side=3, line=1.6, font=2)
mtext("Depth (cm)", side=2, line=1.6, font=2)

lm3 <- lm(log10(Gd)~Depth_mean + Depth2 + Depth3, data=data0)
slm3 <- summary(lm3)
depfz <- data.frame(Depth_mean = seq(2.5,70,2.5),
                    Depth2 = seq(2.5,70,2.5)^2,
                    Depth3 = seq(2.5,70,2.5)^3)
predz <- as.data.frame(predict(lm3, newdata = depfz, interval="confidence"))
predz <- cbind(depfz, predz)
polygon(c(predz$lwr, rev(predz$upr), predz$lwr[1]),
        c(predz$Depth_mean, rev(predz$Depth_mean),predz$Depth_mean[1]),
        col="#ff000040", border="transparent")
lines(predz$fit, predz$Depth_mean, col="blue3")
lines(predz$lwr, predz$Depth_mean, col="red3", lty=2)
lines(predz$upr, predz$Depth_mean, col="red3", lty=2)
with(data0, points(log10(Gd), Depth_mean, cex=1.2, pch=19))
text(par("usr")[1],par("usr")[4],col="blue3",
     labels=paste0("\n\n\nR² = ",signif(slm3$r.sq,3), ";\np = ",
                   signif(1-(pf(summary(lm3)$fst, 1, lm3$df.res)[[1]]),3)),
     pos=4)
legend("bottomright", legend=c("Observations","Cubic model","Model 95% CI"),
       pch=c(19,NA,15), pt.cex=c(1.2,NA,2), col=c(1,2,4), lwd=c(NA,1,NA),
       bg=5, box.col=6, inset=c(0.005,0.1))
```

&nbsp;

## References and R Packages

Graves S, Piepho H, Dorai-Raj LSwhfS (2024). `multcompView`: _Visualizations of Paired Comparisons_. [doi:10.32614/CRAN.package.multcompView](https://doi.org/10.32614/CRAN.package.multcompView){target="_blank"}, R package version 0.1-10, [https://CRAN.R-project.org/package=multcompView](https://CRAN.R-project.org/package=multcompView){target="_blank"}.

Mangiafico, Salvatore S. (2024). `rcompanion`: Functions to Support Extension Education Program Evaluation. version 2.4.36. Rutgers Cooperative Extension. New Brunswick, New Jersey. [https://CRAN.R-project.org/package=rcompanion](https://CRAN.R-project.org/package=rcompanion){target="_blank"}

Snow G (2024). `TeachingDemos`: _Demonstrations for Teaching and Learning_. [doi:10.32614/CRAN.package.TeachingDemos](https://doi.org/10.32614/CRAN.package.TeachingDemos){target="_blank"}, R package version 2.13, [https://CRAN.R-project.org/package=TeachingDemos](https://CRAN.R-project.org/package=TeachingDemos){target="_blank"}.
  
