---
title: "Check normality of distributions"
subtitle: "Many variables at once using a programming loop"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: yeti
    code_folding: show
    self_contained: no
    number_sections: no
    smart: no
    toc: true
    toc_depth: 2
    toc_float: true
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r preliminaries, warning=FALSE, message=FALSE, include=FALSE}
library(car)  ### for powerTransform() function
library(flextable)
  set_flextable_defaults(font.family = "Arial", font.size = 10, 
                       theme_fun = "theme_booktabs", padding = 1)
library(officer)
git <- "https://github.com/Ratey-AtUWA/Learn-R-web/raw/main/"
sv2017 <- read.csv(paste0(git, "sv2017_original.csv"))
```

We need to load the **R** package `car`, as it's needed for the `powerTransform()`
function. For this session we load and use the `sv2017` data from Smith's Lake
and Charles Veryard Reserves in 2017.

```{r prelim-notrun, warning=FALSE, message=FALSE, eval=FALSE}
library(car)
git <- "https://github.com/Ratey-AtUWA/Learn-R-web/raw/main/"
sv2017 <- read.csv(paste0(git, "sv2017_original.csv"))
```

It's often a good idea to make a copy of the data, so we don't lose any
information from the original if something goes horribly wrong. We then check 
our data using the `summary()` function:

```{r copy-data, message=FALSE, warning=FALSE, results='hold'}
data0 <- sv2017
summary(data0)
```

The `summary()` shows us that the first numeric variable we want to check the 
distribution for is `pH`. There are other numeric columns before this (`Group`, 
`Easting`, `Northing`,`Longitude`,`Latitude`) but we don't usually tend to check
the distribution for this type of numeric or integer variable, as they are
really sample identifiers.

If you use this code for your own data, change the value of `v0` to your first
numeric variable of interest (*i.e*. the first you want to check the 
distribution of).

```{r}
v0 <- "pH"
```

### Subset the input data from first relevant numeric column to end
```{r subset-input-data, message=FALSE, warning=FALSE}
data0 <- data0[,which(colnames(data0)==v0):ncol(data0)]
```

We then create a temporary object `names.of.cols` storing the names of variables 
to be transformed, and a vector `colz` of the column indices which are numeric:

```{r}
names.of.cols <- names(data0)
colzTF <- unlist(lapply(data0, is.numeric), use.names = FALSE)
colz <- which(colzTF==TRUE)
```

We make an empty output data frame `transf_results` to hold the results of our
distribution checks:

```{r}
transf_results <- data.frame("Variable"=rep(NA,length(colz)),
              "W_orig"=rep(NA,length(colz)),
              "p_orig"=rep(NA,length(colz)), "W_log_tr"=rep(NA,length(colz)),
              "p_log_tr"=rep(NA,length(colz)), "W_pow_tr"=rep(NA,length(colz)),
              "p_pow_tr"=rep(NA,length(colz)), "Pow_term"=rep(NA,length(colz)))
```

We then start a programming loop using `for(...)` that assesses variable
distributions and writes the statistical output to the data frame
`transf_results` we created earlier.

Note that we create R objects to hold the results of statistical tests &ndash; 
this stores them temporarily so that we can write them to the results data frame.
<br />
The object `pt1` that hold the results of the `powerTansform()` function 
contains an item `pt1$lambda` which is the relevant power term for each
variable. Each time we cycle through the `for()` loop, we write a row of the
output data frame `transf_results`. We use `signif()` to round to a specified
number (`4`) of significant digits.

```{r transformations-loop, message=FALSE, warning=FALSE, paged.print=FALSE}
for (i in colz) {
  pt1 <- powerTransform(data0[, i])
  sw0 <- shapiro.test(data0[, i])
  sw1 <- shapiro.test(log10(data0[, i]))
  sw2 <- shapiro.test((data0[, i]) ^ as.numeric(pt1$lambda))
  transf_results[i-(colz[1]-1),] <- c(names.of.cols[i], signif(sw0$statistic, 4),
                                 signif(sw0$p.value, 4), signif(sw1$statistic, 4),
                                 signif(sw1$p.value, 4), signif(sw2$statistic, 4),
                                 signif(sw2$p.value, 4), signif(as.vector(pt1$lambda), 4))
}
# output to console (screen)
{cat("Table 1: Shapiro-Wilk statistics and p-values for untransformed (_orig) and transformed
(_log, _pow) variables from soil and sediemnt analysis at Smith's Lake Reserve.\n\n")
print(transf_results, row.names = FALSE)}
```

We can export our results to a csv file for Excel (if desired):
```{r expot, eval=FALSE}
write.csv(transf_results, file = "transformations.csv", row.names = FALSE)
```

Alternatively, we can make a publication-quality table using the `flextable`
**R** package (Table 1).

```{r}
flextable(transf_results) |> bold(part="header") |> width(width=2, unit="cm") |>
  set_caption(caption="Table 1: Table summarising normality tests for distributions of variables when untransformed, log10-transformed, or power-transformed. [W_orig = Shapiro-Wilk statistic (W) for the untransfomed variable; p_orig = Shapiro-Wilk p-value for the untransfomed variable; W_log_tr = W for log10-transformed; p_log_tr = p-value for log10-transformed; W_pow_tr = W for power-transformed; p_pow_tr = p-value for power-transformed; Pow_term = exponent for power transformation.] Data from Smiths-Veryard 2017.", align_with_table=F, fp_p=fp_par(text.align = "left", padding.bottom = 6))
```

```{r remove temporary objects, message=FALSE, warning=FALSE, include=FALSE}
### to keep R workspace tidy
rm(list=c("c1","cn","S_W_raw","names.of.cols","pt1","sw0","sw1","sw2","i"))
```

<p>&nbsp;</p>
