---
title: "Use geotagged jpeg images as map backgrounds"
subtitle: "Maps with better aerial photographs"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document:
    highlight: kate
---

<style type="text/css">
  body{
  font-size: 11pt;
}
</style>

# Introduction

Sometimes we can obtain the best map background images in the form of
georeferenced images, such as geoTIFF or jpeg. In this example we'll use a jpeg
file downloaded from the excellent
[NearMap](https://www.nearmap.com/au/en){target="_blank"} site (my university
has a subscription 😊). Specialist mapping websites may have better resolution,
or have aerial photographs from the most appropriate time of year for our
mapping exercise.

First we load the packages that we need:

```{r load-packages, message=FALSE, warning=FALSE, echo=1:5}
library(sf)
library(raster)
library(viridis)
library(stars)
library(prettymapr)
# library(rgdal)
```

We should note that the `jpeg` file itself doesn't have spatial information, even 
though we may have downloaded it as a georeferenced image. With jpeg images, 
georeferencing comes in the form of a companion ('sidecar') file, with the same
name but having the `.jgw` extension. We can read this directly, but the
functions we will use below will look for this file and use the spatial
information directly. The `.jgw` file has 6 lines:

1. length of a pixel in the x direction (horizontal)
2. angle of rotation (is usually 0 or ignored)
3. angle of rotation (is usually 0 or ignored)
4. negative length of a pixel in the y direction (vertical)
5. x coordinate at centre of pixel in the top left corner of the image
6. y coordinate at centre of pixel in the top left corner of the image

So long as the file names of the `.jpg` and `.jpw` files match, the raster
extent and coordinates are obtained or calculated from the values in the `.jpw`
file.

# Reading the raster spatial information in the georeferenced jpeg image

We use the `stars` package, which plays nicely with the `sf` package, to do most
of the work (the `sf` package was introduced on [another page on maps in
**R**](Rmaps.html). The `read_stars()` function will find and use the
information in the `.jpw` file without us specifying that file explicitly.

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
(AF_nearmap <- 
  read_stars("EPSG32750_Date20230226_Lat-31.918263_Lon115.944872_Mpp0.597.jpg"))
```

We notice that, apart from the long and informative filename, the resulting 
`stars` object does not have a coordinate reference system included, so we need 
to add the appropriate one manually:

```{r}
st_crs(AF_nearmap) <- st_crs(32750)
strtrim(st_crs(AF_nearmap),30)
```

The messy output above shows that we have what we expect: UTM Zone 50 (south),
based on the [WGS84 datum](https://support.virtual-surveyor.com/en/support/solutions/articles/1000261351-what-is-wgs84-){target="_blank"}.

From the summary of the `read_stars()` output above (*i.e*. `band    1    3`), 
we can see that the image has three bands of information, corresponding to the
red, green and blue colour channels in the jpeg image.

To plot as a combined colour image we need to merge these 3 channels into in 
single data layer of colour information; fortunately, there is a function 
for this in the `stars` package: `st_rgb()`:

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
AF_rgb <- st_rgb(AF_nearmap)
print(AF_rgb)
```

We need some data to plot, so let's read some

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
git <- "https://github.com/Ratey-AtUWA/cybloRg/raw/main/"
afs1922 <- read.csv(paste0(git,"afs1922edit.csv"), stringsAsFactors = TRUE)
afs1922$Year <- as.factor(afs1922$Year)
(afsREE <- st_as_sf(x = afs1922[,c("Year","Easting", "Northing","Al","Fe","REE")],
         coords = c("Easting", "Northing"), crs = st_crs(32750)))
```

Now we can plot a map with some data on it! We use the `image()` function to
plot the stars raster; we could also use `plot()` if we specify the argument
`reset = FALSE`.

```{r afr-nearmap, fig.align='center', fig.height=6.6, fig.width=10.02, message=FALSE, warning=FALSE, fig.cap="Figure 1: Map of sampling locations at Ashfield Flats Reserve from 2019-2022 plotted on a Nearmap spatial raster.", results='hold'}
palette(c("white", mako(8, dir=-1, alpha=0.7), "black","transparent")) # custom palette
par(mar=c(3,3,1,1), oma=c(0,0,0,0), mgp=c(1.6,0.3,0), tcl=-0.2, 
    font.lab=2, cex.lab=1.2, lend="square", ljoin="mitre")
# Plot an empty frame, same size as the stars object
plot(matrix(st_bbox(AF_rgb), nrow=2, byrow=T), asp=1, type="n", 
     xaxs="i", yaxs="i",                      # no space around data range
     xlab="Easting (m)", ylab="Northing (m)") 
image(AF_rgb, add=TRUE) # add the georeferenced image, and
box()                   # re-draw the box in case of overlap
# plot the points, symbols varying by Year:
#   1. first black unfilled to create 'shadow' behind for better contrast... 
points(st_coordinates(afsREE), pch=c(1,0,5,2)[afsREE$Year], col=10, lwd=2,
       cex=c(1.7,1.6,1.6,1.6)[afsREE$Year])
#   2.  then overplot  with white border. 
points(st_coordinates(afsREE), pch=c(21:24)[afsREE$Year], col=1, lwd=1.5,
       bg=c(2,4,6,8)[afsREE$Year], cex=c(1.5,1.4,1.4,1.4)[afsREE$Year])
# double plot north arrow with small offset to create shadow behind
addnortharrow(pos="topright",padin = c(0.22,0.22), 
              text.col = 10, border = 10, scale=1.4)
addnortharrow(pos="topright",padin = c(0.2,0.2), 
              text.col = 1, border = 1, scale=1.4)
addscalebar(plotepsg=32750, pos="bottomleft",linecol = 1,label.col = 1,
            padin = c(0.22,0.18),htin = 0.2, label.cex = 1.5, widthhint = 0.15)
# add legend for map image and projection information
legend("topleft", title=expression(bold("Ashfield Flats Reserve")),
       legend=c("Photo map source: NearMap Feb 2023 ",
                "EPSG: 32750",
                "(UTM Zone 50 S, WGS84)"),
       box.col="white", bg="#ffffffd0", inset=c(0.01,0.01), 
       title.cex = 1.4, cex=0.9, y.intersp=0.9)
# add legend to identify points by year
legend("bottomright", title=expression(bold("Year")),
       legend=levels(afsREE$Year), text.col = 1, pch=c(21,22,23,24), 
       pt.bg=c(2,4,6,8), col=1, box.col=1, pt.lwd=2, pt.cex=c(1.6,1.4,1.4,1.4), 
       bg = "#707062",inset=c(0.03,0.1), cex=1.2, y.intersp = 1)
```

<p>&nbsp;</p>

For comparison we can also use the `add.geom =` argument in the `image()` or
`plot()` function as an alternative way to plot the points.

```{r afr-nearmap-addgeom, fig.align='center', fig.height=6.6, fig.width=10.02, message=FALSE, warning=FALSE, fig.cap="Figure 2: Map of sampling locations at Ashfield Flats Reserve from 2019-2022 plotted on a Nearmap spatial raster using `plot.geom` as an argument to `image()`.", results='hold'}
palette(c("white",plasma(8, alpha=0.7, beg=0.4), "black","transparent"))
par(mar=c(3,3,1,1), oma=c(0,0,0,0), mgp=c(1.7,0.3,0), tcl=0.25,
    lend="square", ljoin="mitre", font.lab=2)
plot(matrix(st_bbox(AF_rgb), nrow=2, byrow=T), asp=1, type="n", 
     xaxs="i", yaxs="i",
     xlab="Easting (m)", ylab="Northing (m)")
# note use of add.geom=... in image() function
image(AF_rgb, add=TRUE, 
      add.geom=list(afsREE, col=10, 
                    pch=c(21:24)[afsREE$Year], 
                    bg=c(3,5,7,9)[afsREE$Year], 
                    cex=c(1.6,1.4,1.4,1.4)[afsREE$Year]))
box()
addnortharrow(pos="topright",padin = c(0.22,0.22), 
              text.col = 10, border = 10, scale=1.4)
addnortharrow(pos="topright",padin = c(0.2,0.2), 
              text.col = 1, border = 1, scale=1.4)
addscalebar(plotepsg=32750, pos="bottomleft",linecol = 1,label.col = 1,
            padin = c(0.22,0.18),htin = 0.2, label.cex = 1.5, widthhint = 0.15)
# add legend for map image and projection information
legend("topleft", title=expression(bold("Ashfield Flats Reserve")),
       legend=c("Photo map source: NearMap Feb 2023 ",
                "EPSG: 32750",
                "(UTM Zone 50 S, WGS84)"),
       box.col="white", bg="#ffffffd0", inset=c(0.01,0.01), 
       title.cex = 1.4, cex=0.9, y.intersp=0.9)
# add legend to identify points by year
legend("bottomright", title=expression(bold("Year")),
       legend=levels(afsREE$Year), text.col = 1, pch=c(21,22,23,24), 
       pt.bg=c(3,5,7,9), col=10, box.col=1, pt.lwd=1, pt.cex=c(1.6,1.4,1.4,1.4), 
       bg = "#606054c0",inset=c(0.03,0.1), cex=1.2, y.intersp = 1)
```

<p>&nbsp;</p>

The result is similar to adding points separately using `points()`, but we have
somewhat less control over plot appearance. We could also add our data in a way
that shows values of a variable, as described in the [page on maps in
**R**](Rmaps.html).

## Packages

Bivand R, Keitt T, Rowlingson B (2022). *rgdal: Bindings for the 'Geospatial' Data Abstraction*
  *Library*. R package version 1.6-2, [https://CRAN.R-project.org/package=rgdal](https://CRAN.R-project.org/package=rgdal){target="_blank"}.
  
Dunnington D (2022). _prettymapr: Scale Bar, North Arrow, and Pretty Margins in
R_. R package version 0.2.4,
[https://CRAN.R-project.org/package=prettymapr](https://CRAN.R-project.org/package=prettymapr){target="_blank"}.
  
Garnier S., Ross N., Rudis R., Camargo A.P., Sciaini M., and Scherer C.
  (2021). *Rvision - Colorblind-Friendly Color Maps for R*. R package version 0.6.2. [https://sjmgarnier.github.io/viridis/](https://sjmgarnier.github.io/viridis/){target="_blank"}

Hijmans R (2022). _raster: Geographic Data Analysis and Modeling_. R package
version 3.6-11,
[https://CRAN.R-project.org/package=raster](https://CRAN.R-project.org/package=raster){target="_blank"}.

Pebesma, E., 2018. Simple Features for R: Standardized Support for Spatial Vector Data. *The R*
  *Journal* **10** (1), 439-446, [https://doi.org/10.32614/RJ-2018-009](https://doi.org/10.32614/RJ-2018-009){target="_blank"}

Pebesma E (2022). _stars: Spatiotemporal Arrays, Raster and Vector Data Cubes_.
R package version 0.6-0,
[https://CRAN.R-project.org/package=stars](https://CRAN.R-project.org/package=stars){target="_blank"}.

