---
title: "Statistical Relationships"
subtitle: "Grouped linear regression"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: html_document
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages-etc, include=FALSE}
library(car)
library(flextable)
library(officer)
library(viridis)
library(png)
library(car)
library(lmtest)
library(effects)

set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 10, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)

addImg <- function(obj, x = NULL, y = NULL, width = NULL, interpolate = TRUE){
  if(is.null(x) | is.null(y) | is.null(width)){stop("Must provide args 'x', 'y', and 'width'")}
  USR <- par()$usr ; PIN <- par()$pin ; DIM <- dim(obj) ; ARp <- DIM[1]/DIM[2]
  WIDi <- width/(USR[2]-USR[1])*PIN[1] ;   HEIi <- WIDi * ARp 
  HEIu <- HEIi/PIN[2]*(USR[4]-USR[3]) 
  rasterImage(image = obj, xleft = x-(width/2), xright = x+(width/2),
            ybottom = y-(HEIu/2), ytop = y+(HEIu/2), interpolate = interpolate)
}
```

<div style="border: 2px solid #039; padding: 8px;">
<p style="text-align:center; font-size:12pt;">
<em>Relationships pages</em>: [Correlations](correl.html){style="color:#04b;"} | 
[Simple linear regression](regression.html){style="color:#04b;"} | 
[Grouped linear regression](reg-group.html){style="color:#04b;"} | 
[Multiple linear regression](reg-multi.html){style="color:#04b;"}</p>
</div>

&nbsp;

## Introduction

In a [previous session on simple regression](regression.html), we finished by 
suggesting that one way we could make regression prediction more accurate was by 

> allowing regression relationships to differ for different groups in our data 
> (defined by a factor variable) &ndash; *regression by groups*

This session will continue or efforts to predict Cr from Al, but we know the
data were derived from analysis of samples from diverse types of environment.
For that reason, it seems reasonable to evaluate a regression model which allows
the regression coefficients (intercept and slope) to differ for different sample
types.

<div style="border: 2px solid #039; background-color:#e8e8e8; padding: 8px;">
Using the R code file provided and the `afs1923` dataset:

- Run a grouped regression model,
- Assess whether or not the grouped regression model meets the statistical assumptions,
- Evaluate whether a grouped regression model is a statistically significant improvement on the comparable ungrouped model, and
- Make sure you understand the concepts for all the statistical methods you use, and can interpret all the output!

<span style="font-size: 12pt;"><iframe width="360" height="240" src="https://echo360.net.au/media/f444fd3e-9d67-468b-aeb2-c2b27ca26c5e/public?autoplay=false&amp;automute=false" title="E3361+Correl+Regr.mp4" allowfullscreen="allowfullscreen" webkitallowfullscreen="webkitallowfullscreen" mozallowfullscreen="mozallowfullscreen"></iframe></span>

<p><a href="https://raw.githubusercontent.com/Ratey-AtUWA/Learn-R/main/afs1923.csv" target="_blank"><span style="font-size: 12pt;">💾&nbsp;The afs1923 dataset in csv (comma-separated) format</span></a></p>
<p><a href="https://lms.uwa.edu.au/bbcswebdav/pid-3088955-dt-content-rid-40243114_1/xid-40243114_1" target="_blank"><span style="font-size: 12pt;">🔣&nbsp;R code file for this and related workshop sessions</span></a></p>
</div>

<p>&nbsp;</p> 

As previously, our first step is to read the data, and change anything we need to:

```{r read-data, message=FALSE, warning=FALSE}
git <- "https://github.com/Ratey-AtUWA/Learn-R/raw/main/"
afs1923 <- read.csv(paste0(git,"afs1923.csv"), stringsAsFactors = TRUE)
# re-order the factor levels
afs1923$Type <- factor(afs1923$Type, levels=c("Drain_Sed","Lake_Sed","Saltmarsh","Other"))
afs1923$Type2 <- 
  factor(afs1923$Type2, 
         levels=c("Drain_Sed","Lake_Sed","Saltmarsh W","Saltmarsh E","Other"))
```

Next, we inspect a scatterplot of the variables which matches the kind of 
regression model we intend to apply (Figure 1):

```{r sp-Cr-Al-byType, warning=FALSE, message=FALSE, results='hold', fig.width=5, fig.height=4, fig.cap="Figure 1: Different relationships predicting Cr from Al depending on the sample type (data from analysis of sediments from Ashfield Flats, 2019-2023."}
library(car)
par(mar=c(3,3,0.5,0.5), mgp=c(1.3,0.2,0), tcl=0.2, font.lab=2)
scatterplot(Cr ~ Al | Type2, data=afs1923, smooth=FALSE,
            col=inferno(nlevels(afs1923$Type2), end=0.75), pch=c(5,17,15,19,10), 
            legend=list(coords="bottomright", title="Sample type"))
```

## Predicting chromium from aluminium using a regression model which varies by groups 

Note the syntax used to separate by factor categories:

```{r grouped-1predictor-regmodel, results='hold'}
lmCrAl_byType <- lm(Cr ~ Al * Type2, data=afs1923)
summary(lmCrAl_byType)
```

This is similar output to simple linear regression in the previous example, but
the `Coefficients:` table is much more complicated.

- The first two rows of the `Coefficients:` table under the headings give the
intercept and slope for the 'base case', which by default is the first group in
the factor separating the groups. In this example the first level of the factor
`Type` is `Drain_Sed` (frustratingly, the output does not show this). 
- The next 4 rows, starting with the factor name (*i.e*. `Type2Lake_Sed`, *etc*.),
show the *difference* between the **intercepts** for `Lake_Sed`, `Saltmarsh W`,
`Saltmarsh E`, and `Other` groups compared with the base case. 
- Similarly, the final rows, beginning with the predictor variable name
(`Al:Type2Lake_Sed`, `Al:Type2Saltmarsh W`, `Al:Type2Saltmarsh E`, and
`Al:Type2Other`), show the *difference* between the **slopes** for `Lake_Sed`,
`Saltmarsh W`, `Saltmarsh E`, and `Other` groups compared with the base case
(shown as `Al` but actually the slope for the `Drain_Sed` group).

## Compare the two models 

Sometimes models can have greater R^2^ but only because we've made them more
complex by grouping our observations or by adding more predictors. 
**We want the simplest model possible**. We compare the models using an analysis 
of variance with the `anova()` function (where the null hypothesis is equal
predictive ability). The models compared need to be **nested**, that is, one is
a subset of the other.

Before we do this, we need to re-create the simple model from a 
[previous session](regression.html):

```{r}
lmCrAlsimple <- lm(Cr ~ Al, data=afs1923)
summary(lmCrAlsimple)
```

So our more complex (grouped) linear model has greater R^2^ (0.836) than the 
simple linear model (R^2^&nbsp;=&nbsp;0.791). We now perform the `anova` test to 
see whether this improvement is significant:

```{r}
anova(lmCrAlsimple, lmCrAl_byType)
```

The p-value (`Pr(>F)`) is &sime;&nbsp;4.4&times;10^&minus;14^, *i.e*. below 0.05, 
so we reject H~0~ that the models have equivalent predictive ability. In that 
case the model which explains more variance in the dependent variable `Cr` is 
best; that is, the *grouped linear model* with R^2^&nbsp;=&nbsp;0.836 (*i.e*.,
explaining 83.6% of the variance in `Cr` based on the variance in `Al` divided 
into groups by the factor `Type2`).

<p id="Occam">The choice of models should be based on an interpretation of 
&ldquo;*Occam's Razor*&rdquo; a.k.a. the &ldquo;*Principle of Parsimony*&rdquo;:</p>

>  &ldquo;Plurality must never be posited without necessity.&rdquo;  
> &mdash; *William of Ockham (also spelled Occam) (1285 - 1348), English theologian, logician, and Franciscan friar*

In practice this means that if more than one hypothesis or model explains our 
data equally well, we should always choose the simplest one.

Another way we can test for the most appropriate model is by using the AIC or
*Aikake Information Criterion*, which &lsquo;penalises&rsquo; a model having 
more parameters. A better model will have a lower value of AIC, and the built-in
penalty means that for equivalent models (i.e. non-significant ANOVA) the model
with more parameters will have greater AIC, and so we should select the simpler
model with the least AIC value. For our two models, AIC shoud still be lower for
the more complex model (otherwise we are in trouble, since the ANOVA and AIC
would disagree, and then which one would we choose?). Anyway, let's find out:

```{r compare-simple-grouped-AIC, message=FALSE, warning=FALSE, paged.print=FALSE}
AIC(lmCrAlsimple, lmCrAl_byType)
```

The output from the `AIC()` function shows that even though the degrees of
freedom `df` for the grouped model (11) is much greater than for the simple
model (3), the grouped model still has a lower AIC value (&sime;2392) than the
simple model (&sime;2456). So, the ANOVA and AIC do agree after all 😌.

We can see a little more detail about the individual group-wise relationships or 
&lsquo;effects&rsquo; by plotting an effects plot using functions from the 
`effects` package (Figure 2):

```{r plot-grouped-effects, message=FALSE, warning=FALSE, fig.height=5, fig.width=6, fig.cap="Figure 2: Effect plots showing the effects of interacting predictors `Al * Type2` on Cr, using data from sediments at Ashfield Flats 2019-2023.", results='hold'}
library(effects)
plot(allEffects(lmCrAl_byType))
```

So even though the plots in Figure 1 and Figure 2 suggest that each group has a 
similar `Cr ~ Al` relationship, the slopes and intercepts differ enough between
groups to give an improved prediction over the simple model.

## Regression assumptions

The assumptions implicit in linear regression were presented in the 
[session on simple linear regression models](regression.html#assumptions). We run the same 
set of diagnostic tests on more complex regression models as well.

### Diagnostic plots for Grouped linear regression, `Cr ~ Al * Type2`

```{r diag-plots-lm-grouped, fig.height=5, fig.width=5, fig.cap="Diagnostic plots for a grouped linear model predicting Cr from Al with obervations grouped by sample type.", message=FALSE, warning=FALSE, echo=1:2}
par(mfrow=c(2,2), mar=c(3,3,2,1), mgp=c(1.3,0.2,0), tcl=0.2, font.lab=2)
plot(lmCrAl_byType, col="#003087")
par(mfrow=c(1,1))
```

## Formal statistical tests of regression assumptions

As with simple linear regression, we can test all of the assumptions with formal
statistical tests (see the [simple linear regression session](regression.html#diagtests)
for explanations and null hypotheses).

```{r shapiro-test-resids, results='hold'}
shapiro.test(lmCrAl_byType$residuals) # normality
```

```{r bgtest, results='hold'}
require(lmtest)
bgtest(lmCrAl_byType) # autocorrelation (independence)
```

```{r bptest, results='hold'}
bptest(lmCrAl_byType) # homoscedasticity
```

```{r rainbow-test, results='hold'}
raintest(lmCrAl_byType) # linearity
```

```{r outlier-test, results='hold', paged.print=FALSE}
require(car)
outlierTest(lmCrAl_byType) # influential observations (outliers)
```

```{r cokks-dist, results='hold', paged.print=FALSE}
summary(cooks.distance(lmCrAl_byType)) # influential observations
```

```{r influence-measures, results='hide'}
infl.lmCrAl <- influence.measures(lmCrAl_byType)
ninf <- nrow(summary(infl.lmCrAl)) # number of possibly influential observations
modform <- deparse(lmCrAl_byType$call)
rninf <- row.names(summary(infl.lmCrAl)) # IDs of possibly influential observations
# if you run this code chunk yourself you will see the whole summary(infl.lmCrAl) output

```
```{r summary-infl-meas, results='hold'}
cat(paste0("For the model '", modform, "' there are ", ninf, 
           " potentially influential observations.\n\n"))
cat("The", ninf, "rows containing potentially influential obervations are:\n"); as.integer(rninf)
```

To summarise:

```{r diagnost-summ, echo=FALSE}
diagsumm <- data.frame(Test=c("Shapiro-Wilk", "Breusch-Godfrey","Breusch-Pagan","Rainbow","Outlier","Cook's Distance","Infuence Measures"),
                       Assessing=c("Are residuals normally distributed?", "Are residuals independent?","Are residuals homoscedastic?","Is the relationship linear?","Are there significant outliers (Bonferroni)?","Are there any potentially influential observations?","Are there any potentially influential observations?"),
                       Results=c("No (qq-plot suggests long tails)", "No, there is autocorrelation","No (p<0.05)","Yes (p>0.05)","2 points are possible outliers","No (no values > 0.5)","Yes (36 out of 335 observations)"))
flextable(diagsumm) |> width(width=c(4,8,7), unit = "cm") |> valign(valign = "top") |> padding(padding=2) |> fontsize(size=10, part="all") |> 
  set_caption(caption="Table 1: Summary of regression diagnostic tests for a grouped linear model predicting Cr from Al (grouped by Type) in sediments from Ashfield Flats Reserve, based on data from samples collected from 2019-2023.", align_with_table=F, fp_p=fp_par(text.align = "left", padding.bottom = 6))
```

<p>&nbsp;</p>

The summary of the diagnostic tests in Table 1 shows that not all the formal
assumptions of regression are fulfilled by our model. Even though the model
may not be strictly statistically valid, the regression output showed that 
&sime;&nbsp;84% of the variance in Cr could be explained by Al when observations
were grouped by sample type. So, from a practical perspective, this grouped
regression model could also be used for prediction, and the greater R^2^ makes 
it a better prediction option than the simple model discussed in 
[another session](regression.html).

The next option would be a model which has more than one predictor variable for our dependent variable &ndash; [multiple regression](reg-multi.html).
</div>

<p>&nbsp;</p>

# References

Fox, J. and Weisberg, S. (2019). *An {R} Companion to Applied Regression, Third Edition*. Thousand Oaks CA: Sage. [https://socialsciences.mcmaster.ca/jfox/Books/Companion/](https://socialsciences.mcmaster.ca/jfox/Books/Companion/){target="_blank"} (**R** package `car`)

Garnier S, Ross N, Rudis R, Camargo AP, Sciaini M, Scherer C (2021). *Rvision &ndash; Colorblind-Friendly Color Maps for R*. R package version 0.6.2. (**R** package `viridis`)

Reimann, C., Filzmoser, P., Garrett, R.G., Dutter, R., (2008). *Statistical Data Analysis Explained: Applied Environmental Statistics with R*. John Wiley & Sons, Chichester, England (see Chapter 16).

Zeileis, A. and Hothorn, T. (2002). Diagnostic Checking in Regression Relationships. *R News* **2**(3), 7-10. [https://CRAN.R-project.org/doc/Rnews/](https://CRAN.R-project.org/doc/Rnews/){target="_blank"} (**R** package `lmtest`)

<p>&nbsp;</p>
