---
title: "Time Series Analysis"
subtitle: "Concepts using soil temperature data"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    self_contained: no
    number_sections: no
    smart: no
    toc: true
    toc_depth: 3
    toc_float: true
---

<div style="border: 2px solid #039; background-color:#e8e8e8; padding: 8px;">
<p><span style="font-size:14pt;">This guide is organised into 3 main sections:</span></p>
<ol>
<li><span style="font-size:14pt;"><a href="#part1">Setting up R to analyse time series data, and the associated data formats</a>;</span></li>
<li><span style="font-size:14pt;"><a href="#part2">Understanding the concepts, components and features of time series data</a>;</span></li>
<li><span style="font-size:14pt;"><a href="#part3">Forecasting time series using basic trend analysis and more complex ARIMA (and other) models</a> &ndash; the type of content we would include in reports.</span></li>
</ol>
</div>

&nbsp;

## Set up the R environment for time series analysis

We will need the additional functions in several **R** packages for
specialized time series and other functions in R...

```{r setup-run, message=FALSE, warning=FALSE, include=TRUE, results='hide'}
# Load the packages we need 
library(zoo)      # for basic irregular time series functions
library(xts)      # we need the xts "eXtended Time Series" format for some functions
library(Kendall)  # for trend analysis with Mann-Kendall test
library(trend)    # for trend analysis using the Sen slope
library(forecast) # for time series forecasting with ARIMA and exponential smoothing 
library(tseries)  # for assessing stationarity using Augmented Dickey-Fuller test
library(lmtest)   # for Breusch-Pagan heteroscedasticity test etc.
library(car)      # for various commonly-used functions
library(ggplot2)  # alternative to base R plots

# (optional) make a better colour palette than the R default!
# still not ideal for colourblindness (consider using viridis package)
palette(c("black","purple4","#0041B2","darkslategray4",
          "gold3","orangered","darkred","dimgray"))
par(mfrow=c(3,1), mar=c(4,4,1,1), mgp=c(1.7,0.3,0), tcl=0.25, font.lab=2)
```

<p id="part1">&nbsp;</p>

# 1. How R reads and stores time series data

We first read the data into a `data frame` &ndash; this is how R often
stores data &ndash; it's not the format we need, but we'll use it for
comparison.

## Non-time series object for comparison

```{r read-into-dataframe, message=FALSE, warning=FALSE, results='hold'}
git <- "https://raw.githubusercontent.com/Ratey-AtUWA/Learn-R-web/main/"
soiltemp <- read.csv(paste0(git,"soiltemp2.csv"))
colnames(soiltemp) <- c("Date","temp")
```

&nbsp;

### Do some checks of the (non-time series) data

```{r check-dataframe, message=FALSE, warning=FALSE, results='hold'}
summary(soiltemp) # simple summary of each column
str(soiltemp) # more detailed information about the R object 
# ('str'=structure, i.e. how the data are organised in the R object)
```

&nbsp;

We can also check the general form of the data with a plot (Figure 1). 
From this plot and from the text summary output above, we can see that 
the `Date` column is not a date or time, or even numeric.

```{r plot-df, fig.align='center', fig.cap="Figure 1: Plot of soil temperature time series data which is not (yet) formatted as a time series.", fig.height=3.5, fig.width=7, message=FALSE, warning=FALSE, out.width="65%", results='hold'}
par(mfrow = c(1,1), mar = c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, oma = c(0,0,0,0), font.lab=2)
with(soiltemp, plot(temp, type = "l", col = 7))
```

&nbsp;

## Reading and storing data as time series

We really want the data in a different type
of R object! We use the `read.csv.zoo()` function from the `zoo`
package to read the data in as a **time series**. This gives a data
object of class `zoo` which is much more flexible than the default
time series object in R.

We need to specify a format that our date/time information is stored 
as. The character string `"%Y-%m-%d %H:%M:%S"` looks unusual but it 
should make sense (`%Y` is 4-digit year, and so on). If we have opened 
and saved our `.csv` file in Excel, the format will probably change,
so we should try `format = "%d/%m/%Y %H:%M:%S"`. If in doubt, try to match the format win the actual `.csv` file.

```{r read-summary-csv2zoo, message=FALSE, warning=FALSE, results='hold'}
soiltemp_T15_zoo <- read.csv.zoo(paste0(git,"soiltemp2.csv"),
                               format = "%Y-%m-%d %H:%M:%S", 
                               tz = "Australia/Perth", 
                               index.column=1,
                               header = TRUE)
# do some quick checks of the new R object:
summary(soiltemp_T15_zoo) ; cat("\n")
str(soiltemp_T15_zoo) # POSIXct in the output is a date-time format
```

We can see that the column `Date` has now become `Index`, and it's 
correctly stored as date/time information, in a format called 
`POSIXct` (for more information, run `?DateTimeClasses` in the
Console. Again, it's usually useful to check our data with a plot
(Figure 2).

```{r plot-zoo, fig.align='center', fig.cap="Figure 2: Plot of soil temperature time series data which is formatted as a zoo time series object.", fig.height=3.5, fig.width=7, out.width="65%", results='hold'}
par(mfrow=c(1,1), mar=c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, font.lab=2)
plot(soiltemp_T15_zoo, col = 3) 
```

&nbsp;

Sometimes we need to use another time series data format, `xts`
(e<u>X</u>tended <u>T</u>ime <u>S</u>eries), which allows us to use
more functions...

&nbsp;

### Make an xts object from our zoo object

```{r make xts from zoo and check}
soiltemp_T15_xts <- as.xts(soiltemp_T15_zoo)
str(soiltemp_T15_xts) # just to check
```

<p id="part2">&nbsp;</p>

# 2. Exploratory data analysis and decomposition of time series

This section is designed to help you understand the features and components of time series, and some of the statistical methods that we use to analyse them. We also look at time series decomposition. The important issues are:

- Do we need to **transform** our time series variable?
- Is our time series **stationary**?
- Can we separate out a **trend** in out variable *vs*. time?
- Is our time series **seasonal** (periodic) and, if so, what is the periodic frequency?
- Is there significant **autocorrelation** in our data?
- What is left **unexplained** after we isolate the trend and periodicity?

Not all of these would need to be presented in a Report, but it's 
useful for us to understand time series concepts. This is especially 
true for **autocorrelation** and **moving averages**, which both rely 
[at least partially] on past observations to estimate current 
observations. From there it follows that we may be able to use past 
observations to predict **future** observations &ndash; *i.e*., 
**forecasting**.

We first examine a plot of the data in our time series object (the plot for the `xts` time series object is in Figure 3).

```{r plot-xts, fig.align='center', fig.cap="Figure 3: Plot of soil temperature time series data which is formatted as an `xts` time series object.", fig.height=3.5, fig.width=7, message=FALSE, warning=FALSE, out.width="65%", results='hold'}
par(mfrow=c(1,1), mar=c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, font.lab=2)
plot(soiltemp_T15_xts, col = 3, ylab = "Soil temperature (\u00B0C)") # just to check
```

&nbsp;

Plotting the time series variable (with time as the independent
variable!) can give us some initial clues about the overall trend in
the data (in Figure 3, a negative overall slope) and if there is any
periodicity (there do seem to be regular increases and decreases in
soil temperature). The plot in Figure 3 also shows that plotting an
`xts` time series object gives a somewhat more detailed plot than for
a `zoo` formatted time series object (Figure 2).

We often want to inspect plots of both the raw data and using common
transformations (Figure 4). We compare with transformed time series
data &ndash; we may need to do this to meet later modelling
assumptions.

First we change the default plotting parameters using `par(...)`, then
plot the object and some transformed versions (with custom axis
labels).

```{r plot-transf, echo=-1, fig.height=7, fig.width=7, out.width="65%", fig.align='center', fig.cap="Figure 4: Plots of soil temperature time series data as untransformed, log₁₀-transformed, and power transformed values.", results='hold'}
par(mfrow = c(3,1), mar = c(3,3.2,1,1), mgp = c(1.6,0.3,0), tcl = 0.3, font.lab=2)
plot(soiltemp_T15_zoo, ylab = "Temperature (\u00B0C)",
     xlab = "Date", col = 2, lwd = 2, cex.lab = 1.4)
mtext("Untransformed", side=3, line=-1.5, col=2, adj=0.98, font=2)
plot(log10(soiltemp_T15_zoo), 
     ylab = expression(bold(paste(log[10],"(Temperature, \u00B0C)"))),
     xlab = "Date", col = 3, lwd = 2, cex.lab = 1.4)
mtext("log-transformed", side=3, line=-1.5, col=3, adj=0.98, font=2)
pt0 <- powerTransform(coredata(soiltemp_T15_zoo))
if(pt0$lambda<0) {
  plot(
    -1 * (soiltemp_T15_zoo ^ pt0$roundlam), # use rounded power term
    ylab = "power-transf. (Temp., \u00B0C)",
    xlab = "Date",
    col = 7, lwd = 2, cex.lab = 1.4
  )
} else {
  plot((soiltemp_T15_zoo ^ pt0$roundlam), # use rounded power term
       ylab = "power-transf. (Temp., \u00B0C)",
       xlab = "Date",
       col = 7, lwd = 2, cex.lab = 1.4
  )
}
mtext("power-transformed", side=3, line=-1.5, col=7, adj=0.98, font=2)
```

```{r remove-pt0, include=FALSE}
rm(pt0)
```

&nbsp;

> Which of these plots looks like it might be homoscedastic, *i.e*., have constant variance regardless of time?

The R Cookbook (Long and Teetor, 2019) suggests Box-Cox (power)
transforming the variable to "stabilize the variance" (*i.e*. reduce
heteroscedasticity). This does work, but can be better to use
log~10~[variable], as the values are easier to interpret.

There are tests for homoscedasticity, usually based on the residuals 
from fitting a model (*e.g*. the Breusch-Pagan test). For our purposes 
visual assessment is enough.

If we think a transformation is needed, then run something like the
code below (which does a square root transformation, *i.e*.
variable^0.5^). This is just an example &ndash; we may, for example,
decide that the log~10~-transformation is more suitable. If we save 
the results of `powerTransform()` to an object, it's most convenient 
to use the rounded power term (*e.g*. `pt0$roundlam`).

```{r transform if needed not run, eval=FALSE, echo=TRUE}
soiltemp_T15_zoo <- soiltemp_T15_zoo^0.5 #### don't run this chunk of code...
# don't forget the xts version either!
soiltemp_T15_xts <- as.xts(soiltemp_T15_zoo) # ...it's just an example 😊
```

&nbsp;

## Assessing if a time series variable is stationary

A **stationary** variable's mean and variance are not dependent on
time. In other words, for a stationary series, the mean and variance
at any time are representative of the whole series.

If there is a trend for the value of the variable to increase or
decrease, or if there are periodic fluctuations, we don't have a
stationary time series.

Many useful statistical analyses and models for time series models
need a stationary time series as input, or a time series that can be
made stationary with transformations or differencing.

&nbsp;

### Testing for stationarity

```{r adf-tests, message=FALSE, warning=FALSE}
# we need the package 'tseries' for the Augmented Dickey–Fuller (adf) Test
d0 <- adf.test(soiltemp_T15_zoo); print(d0)
d1 <- adf.test(diff(soiltemp_T15_zoo,1)); print(d1)
d2 <- adf.test(diff(diff(soiltemp_T15_zoo,1),1)); print(d2)
```

&nbsp;

### Plot differencing for stationarity

The plots in Figure 5 below show that constant mean and variance
(*i.e*. stationarity) is achieved after a single differencing step.
The `adf.test` p-value suggests that the non-differenced series is
already stationary, but the first sub-plot in Figure 5 does not seem
to support this.

```{r plot-diff, fig.height=8, fig.width=8, out.width="65%", fig.align='center', fig.cap="Figure 5: Plots of soil temperature time series data and its first and second differences, with each sub-plot showing the smoothed (Loess) and linear trends.", results='hold'}
par(mfrow = c(3,1), mar = c(0,4,0,1), oma = c(4,0,1,0), cex.lab = 1.4,
    mgp = c(2.5,0.7,0), font.lab = 2)
plot(soiltemp_T15_zoo, ylab = "Raw data, no differencing",
     xlab="", xaxt="n", col = 8)
lines(loess.smooth(index(soiltemp_T15_zoo),coredata(soiltemp_T15_zoo)), 
      col = 5, lwd = 3)
abline(lm(coredata(soiltemp_T15_zoo) ~ index(soiltemp_T15_zoo)), col = 3)
legend("topright", legend = c("soiltemp_T15_Data", "Loess smoothing","Linear model"),
       cex = 1.8, col = c(8,5,3), lwd = c(1,3,1), bty = "n")
mtext(paste("adf.test p value =",signif(d0$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
plot(diff(soiltemp_T15_zoo,1),
     ylab = "First differencing",
     xlab="", xaxt="n", col = 8)
abline(h = 0, col = "grey", lty = 2)
lines(loess.smooth(index(diff(soiltemp_T15_zoo,1)),coredata(diff(soiltemp_T15_zoo,1))), 
      col = 5, lwd = 3)
abline(lm(coredata(diff(soiltemp_T15_zoo,1)) ~
            index(diff(soiltemp_T15_zoo,1))), col = 3)
mtext(paste("adf.test p value =",signif(d1$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
plot(diff(diff(soiltemp_T15_zoo,1),1),
     ylab = "Second differencing",
     xlab="Date", col = 8)
mtext("Date",side = 1, line = 2.2, font = 2)
abline(h = 0, col = "grey", lty = 2)
lines(loess.smooth(index(diff(diff(soiltemp_T15_zoo,1),1)),
                   coredata(diff(diff(soiltemp_T15_zoo,1),1))), 
      col = 5, lwd = 3)
abline(lm(coredata(diff(diff(soiltemp_T15_zoo,1))) ~ 
            index(diff(diff(soiltemp_T15_zoo,1)))), col = 3)
mtext(paste("adf.test p value =",signif(d2$p.value,3)), 
      side = 1, line = -1.2, adj = 0.05)
```

```{r reset mfrow 1, message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1,1), mar = c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, oma = c(0,0,0,0))
```

&nbsp;

## Finding the trend

### A. Visualising a trend using a moving average

We create a new time series from a moving average for each 24h to
remove daily periodicity using the `rollmean()` function from the
`zoo` package. We know these are hourly data with diurnal fluctuation,
so a rolling mean for chunks of length 24 should be OK, and this seems
to be true based on Figure 6. In some cases the `findfrequency()`
function from the `xts` package can detect the periodic frequency for
us. A moving average will always smooth our data, so a smooth curve
doesn't necessarily mean that we have periodicity (seasonality).

<div style="border: 1px solid #039; background-color:#ffffe0; padding: 8px;">
<span style="font-size:12pt;">**Moving averages** are important parts
of the **ARIMA** family of predictive models for time series, and
that's why we show them here first.</span>

There are possibly better ways of drawing smooth curves to represent our time series data... see below.
</div>

&nbsp;

```{r moving average}
ff <- findfrequency(soiltemp_T15_xts) # often an approximation!! check the data!
cat("Estimated frequency is",ff,"\n")
```

```{r plot-movAv, echo=-1, fig.height=3.5, fig.width=7, fig.align='center', fig.cap="Figure 6: Plots of soil temperature time series data and the 24-hour moving average (ƒ = frequency).", out.width="65%", results='hold'}
par(mfrow = c(1,1), mar = c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, oma = c(0,0,0,0))
soiltemp_T15_movAv <- rollmean(soiltemp_T15_zoo, 24)
plot(soiltemp_T15_zoo, col=8, type="l") # original data
# add the moving average
lines(soiltemp_T15_movAv, lwd = 2, col=2)
legend("topright", bty="n", legend = c("Data","Moving average, \u0192 = 24"), col = c(8,2), lwd = c(1,2))
```

&nbsp;

### B. Showing a trend using a linear (regression) model

This is just to **visualize** the trend. To find if a significant 
trend exists, we use the Mann-Kendall test or Sen's Slope test as
described <a href="#part3">later</a>.

First we create a linear model of the time series...

```{r make linear model}
lm0 <- lm(coredata(soiltemp_T15_zoo) ~ index(soiltemp_T15_zoo))
summary(lm0)
```

...then use a plot (Figure 7) to look at the linear 
relationship.

```{r linear-trend, echo=-1, fig.height=3.5, fig.width=7, out.width="65%", fig.align='center', fig.cap="Figure 7: Plots of soil temperature time series data and the overall trend shown by a linear model.", results='hold'}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
soiltemp_T15_lmfit <- zoo(lm0$fitted.values, index(soiltemp_T15_zoo))
plot(soiltemp_T15_zoo, col = 8, type = "l")
lines(soiltemp_T15_lmfit, col = 2, lty = 2, lwd=2)
legend("topright", bty="n", legend = c("Data","Linear trend"), col = c(8,2), lty = c(1,2), lwd=c(1,2), seg.len=3)
```

&nbsp;

Figure 7 shows that a linear model does not really capture the trend
(even without considering the periodicity) very convincingly. The next
option, 'loess' smoothing, shown below tries to represent the trend 
more accurately.

&nbsp;

### C. Using Locally Estimated Scatterplot Smoothing (loess)

**loess**, sometimes called 'lowess' is a form of locally weighted
non-parametric regression to fit smooth curves to data. The amount of
smoothing is controlled by the `span =` parameter in the
`loess.smooth()` function (from base **R**) &ndash; see the example
using the soil temperature data in Figure 8.

<div style="border: 2px solid #888; padding: 8px;">**NOTE**: loess
smoothing is **not** a standard technique for time series smoothing,
but we use it here for convenience. Loess smoothing does not have the
relationship to ARIMA that moving averaging does, but does allow us to
separate periodicity from random error in time series
decomposition.</div>

```{r loess-trend, echo=-1, fig.height=3.5, fig.width=7, fig.align='center', fig.cap="Figure 8: Plots of soil temperature time series data and the overall trend shown by a loess smoothing model.", out.width="65%", results='hold'}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
y_trend <- loess.smooth(index(soiltemp_T15_zoo), 
                           coredata(soiltemp_T15_zoo), 
                           span = 0.15, evaluation = length(soiltemp_T15_zoo))
plot(soiltemp_T15_zoo, col = 8, type = "l")
soiltemp_T15_trend <- zoo(y_trend$y, index(soiltemp_T15_zoo))
lines(soiltemp_T15_lmfit, col = 2, lty = 2)
lines(soiltemp_T15_trend, col = 3, lwd = 3)
legend("topright", bty = "n", inset = 0.02, cex = 1.25, 
       legend = c("actual data","linear model","loess smoothed"), 
       col = c(8,2,3), lty=c(1,2,1), lwd = c(1,1,3))
```

&nbsp;

## Isolating the Time Series Periodicity

<div style="border: 2px solid #039; background-color:#e8e8e8; padding: 8px;">
<span style="font-size:14pt; color:#600080">**NOTE THAT TIME SERIES DON'T ALWAYS HAVE PERIODICITY !**</span>
</div>

&nbsp;<br />To model the periodicity it is useful to understand the
autocorrelation structure of the time series. We can do this
graphically, first by plotting the autocorrelation function `acf()`
which outputs a plot by default:

```{r acf-plot, echo=-1, fig.height=3, fig.width=6, out.width='65%', fig.align='center', fig.cap="Figure 9: The autocorrelation function plot for soil temperature time series data (the horizontal axis is in units of seconds).", results='hold'}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
acf(soiltemp_T15_xts, main = "")
```

&nbsp;

> What does Figure 9 tell you about autocorrelation in this time series?

Then plot the partial autocorrelation function (`pacf()`)

```{r pacf-plot, echo=-1, fig.height=3, fig.width=6, out.width='65%', fig.align='center', fig.cap="Figure 10: The partial autocorrelation function plot for soil temperature time series data (horizontal axis units are seconds).", results='hold'}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
pacf(soiltemp_T15_xts, main = "")
```

&nbsp;

Interpreting partial autocorrelations (Figure 10) is more complicated
&ndash; refer to [Long & Teetor (2019, *Section 14.15*)](https://rc2e.com/timeseriesanalysis#recipe-id083){target="_blank"}.
Simplistically, partial autocorrelation allows us to identify which
and how many autocorrelations will be needed to model the time series
data.

&nbsp;

## Box-Pierce test for autocorrelation

The null hypothesis H~0~ for the Box-Pierce test (`Box.test()`)is that
no autocorrelation exists at any lag distance (so
p&nbsp;&le;&nbsp;0.05 'rejects' H~0~):

```{r box tests}
Box.test(soiltemp_T15_xts)
Box.test(diff(soiltemp_T15_xts,1)) # 1 difference
Box.test(diff(diff(soiltemp_T15_xts,1),1)) # 2 differences
```

&nbsp;

### Make a time series of the loess residuals

Remember that we made a `loess` model of the time series ... the
residuals (Figure 11) can give us the combination of periodicity
component plus any random variation.

```{r loess-resids, echo=-1, fig.align='center', fig.cap="Figure 11: An incomplete decomposition of the soil temperature time series into smoothed trend and periodicity plus error (noise) components.", fig.height=3.5, fig.width=7, out.width='65%', results='hold'}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
soiltemp_T15_periodic <- soiltemp_T15_zoo - soiltemp_T15_trend
plot(soiltemp_T15_trend, ylim = c(-2,20), lty = 3)
lines(soiltemp_T15_periodic, col = 6, lwd = 2) # just to check
legend("topright", legend = c("LOESS trend", "Periodicity plus noise"), 
       col = c(1,6), lwd = c(1,2), lty = c(3,1))
```

&nbsp;

We can also use *less smoothing* in the loess function to *retain*
periodicity; we adjust the `span =` option (lower values of `span`
give less smoothing; we just need to experiment with different
values). The difference between the data and the less-smoothed loess
should be just 'noise' or '**error**'.

We first generate a loess model which is stored in the object `temp_LOESS2`:

```{r less-smoothed loess model, results='hold'}
temp_LOESS2 <- loess.smooth(index(soiltemp_T15_zoo), 
                           coredata(soiltemp_T15_zoo), 
                           span = 0.012, evaluation = length(soiltemp_T15_zoo))
```

We then use the new loess model to make a time series which contains
both periodic and trend information:

```{r less-smoothed loess time series, results='hold'}
soiltemp_T15_LOESS2 <- zoo(temp_LOESS2$y, index(soiltemp_T15_zoo))
```

The difference between the data and the less-smoothed loess should be
just 'noise' or 'error', so we make a new time series based on this
difference:

```{r error time series, results='hold'}
soiltemp_T15_err <- soiltemp_T15_zoo - soiltemp_T15_LOESS2
```

&nbsp;

Then plot, setting y axis limits to similar scale to original data:

```{r loess-periodic, echo=-1, out.width='65%', fig.align='center', fig.cap="Figure 12: Incomplete decomposition of soil temperature time series data showing the smoothed *trend*, combined *trend + periodicity*, and *error* components.", results='hold'}
par(mar = c(3,3,1,1), mgp = c(1.7,0.3,0), tcl = 0.3, font.lab = 2)
plot(soiltemp_T15_trend, ylim = c(-2,20), lty = 2, lwd = 2, # from above
     xlab = "2014 Date", ylab = "Soil temperature (\u00B0C)")
lines(soiltemp_T15_LOESS2, col = 3, lwd = 2)
lines(soiltemp_T15_err, col = 5) # from a couple of lines above
legend("left", 
       legend = c("Trend (coarse LOESS)", 
                  "Periodicity + trend (fine LOESS)",
                  "Unexplained variation"), 
       col = c(1,3,5), lwd = c(2,2,1), lty = c(2,1,1), bty="n")
```

&nbsp;

We're now in a position to estimate homoscedasticity of our time
series. It's a bit &lsquo;hacky&rsquo;, but we can predict the raw
time series from the smoothed trend+periodicity, then apply the
Breusch-Pagan test to our model (we need the `lmtest` package loaded
earlier for the `bptest()` function):

```{r bptest-time-series, message=FALSE, warning=FALSE, results='hold'}
lm0 <- lm(coredata(soiltemp_T15_zoo) ~ coredata(soiltemp_T15_LOESS2))
bptest(lm0) ; rm(lm0)
```

The `bptest()` p-value is &gt;&nbsp;0.05 so, by this estimate, our 
time series is homoscedastic.

The periodicity by itself should be represented by the difference
between the very smoothed (trend) and less smoothed (trend +
periodicity) loess, as suggested by the two loess curves in Figure 12.
We now make yet another time series object to hold this isolated
periodicity:

```{r make periodic ts}
soiltemp_T15_periodic2 <- soiltemp_T15_LOESS2 - soiltemp_T15_trend
```

&nbsp;

## Show the components of time series decomposition

```{r decomp-plots, fig.height=9, fig.width=8, fig.align='center', fig.cap="Figure 13: Time series decomposition into the three main components (periodicity, trend, and error; the raw data are at the top) for soil temperature (\u00B0C) at 15 cm depth."}
plot(cbind(soiltemp_T15_zoo,soiltemp_T15_periodic2,
     soiltemp_T15_trend, soiltemp_T15_err), main = "", 
     xlab = "Date in 2014", 
     cex.main = 1.5, yax.flip = TRUE, col = c(8,3,4,2), 
     ylim = c(floor(min(soiltemp_T15_periodic2)),
              ceiling(max(soiltemp_T15_zoo))))
x1 <- (0.5*(par("usr")[2]-par("usr")[1]))+par("usr")[1]
y <- (c(0.18, 0.3, 0.6, 0.82)*(par("usr")[4]-par("usr")[3]))+par("usr")[3]
text(rep(x1,4), y, 
     labels = c("Unaccounted variation","Trend","Apparent periodicity","Data"), 
     col = c(2,4,3,8))
```

&nbsp;

Figure 13 shows the original time series data, and the three important
components of the decomposed time series: the *periodicity*, the
*trend*, and the *error* or unexplained variation. We should remember that:

- this time series was relatively easy to decompose into its components, but not all time series will be so 'well-behaved';
- there are different ways in which we can describe the trend and periodicity (we have just used the loess smoothing functions for clear visualization).

<div
style="border: 2px solid #039; background-color:#cff; padding: 8px; text-align: center;">
<span
style="font-family: serif; font-size:12pt;"><strong>This
ends our exploratory data analysis of time series, which leads us into time
series models,<br />including trend and ARIMA forecast modelling.</strong></span>
</div>

&nbsp;

<hr style="height:2px; background-color:#003087;">

<p id="part3">&nbsp;</p>

# 3. Models for Time Series &ndash; Trends and Forecasts

Trends and forecasts of time series are the types of analyses we would 
typically report on (see box below).

## Trends in time series

The simplest model for time series is a trend, *i.e*. an overall
increase or decrease. We don't use the usual correlation or regression
statistics, as we expect autocorrelation in a time series, so the
observations are not independent. We need to use statistical tests
that can handle the autocorrelation dependence between observations,
and these are described below.

<div style="border: 1px solid #039; background-color:#ffffe0; padding: 8px;">
<span style="font-size:13pt;">Simply **finding whether there is a
significant trend, and the direction of that trend** (*i.e*. negative
= decreasing or positive = increasing) may be sufficient from a
regulatory perspective.</span> For example, when monitoring the
attenuation of pollution or checking the effectiveness of remediation,
showing that the contaminant concentrations are decreasing is
important information! (see [Australian Government 2023](https://www.waterquality.gov.au/anz-guidelines/monitoring/data-analysis/data-assess-change){target="_blank"}; 
[Department of Water 2015](https://www.wa.gov.au/government/publications/calculating-trends-nutrient-data){target="_blank"}).
</span></div>

&nbsp;

### (a) Apply the Mann-Kendall test from the 'trend' package

```{r Mann-Kendall test}
mk.test(coredata(soiltemp_T15_zoo))
#    or
# SeasonalMannKendall(soiltemp_T15_zoo) # needs base R regular time series object
```

The output from `mk.test()` shows a negative slope (*i.e*. the value
of `S`), with the `p-value` definitely < 0.05, so we can reject the
null hypothesis of zero slope.

&nbsp;

### (b) Estimate the Sen's slope

```{r Sens slope test}
sens.slope(coredata(soiltemp_T15_zoo))
```

The output from `sens.slope()` also shows a negative slope (*i.e*. the
value of `Sen's slope` at the end of the output block). Again, the
`p-value` is < 0.05, so we can reject the null hypothesis of zero
slope. The `95 percent confidence interval` does not include zero,
also showing that the Sen's slope is significantly negative in this
example.

&nbsp;

## Modelling time series with ARIMA

All the analysis of our data before ARIMA is really exploratory data
analysis of time series:

-   Does our time series have periodicity (seasonality)?
-   Can we get the trend with a moving average?
-   Does our time series have autocorrelation?
-   Can we get stationarity by differencing?

All of these properties are possible components of **ARIMA** models!

&nbsp;<br />[**We recommend using the `xts` format of a time series in
ARIMA model functions and forecasting**.]{style="color: #6000c0;"}

&nbsp;

## Use the `forecast` R package to create an ARIMA model

```{r auto arima}
auto.arima(soiltemp_T15_xts, max.p = 3, max.q = 3, max.d = 0, 
           seasonal = TRUE)
```

The `auto.arima()` function runs a complex algorithm (Hyndman *et al*.
2020) to automatically select the best ARIMA model on the basis of the
**Aikake Information Criterion** (AIC), a statistic which estimates
how much information is 'lost' in the model compared with reality. The
AIC combines how well the model describes the data with a 'penalty'
for increased complexity of the model. Using AIC, a better fitting
model might not be selected if it has too many predictors. The best
ARIMA models will have the **lowest** AIC (or AICc) value.

&nbsp;

## Use output of `auto.arima()` to run arima

The `auto-arima()` algorithm is not perfect! &ndash; but it does
provide a starting point for examining ARIMA models.
The output of the `auto-arima()` function includes a description of
the best model the algorithm found, shown as `ARIMA(p,d,q)`. The
`p`,`d`,`q` are defined in Table 1 below:

```{r echo=FALSE, message=FALSE, warning=FALSE, results='hold'}
arimapars <- data.frame(Parameter=c("p","d","q"),
                        Meaning=c("The number of autoregressive predictors",
                                  "The number of differencing steps",
                                  "The number of moving average predictors"),
                        Info=c("Partial autocorrelation",
                               "Stationarity tests ± differencing",
                               "Stationarity tests ± moving averages"))
library(flextable)
set_flextable_defaults(theme_fun = "theme_zebra", 
                       font.size = 10, fonts_ignore = TRUE)
BorderDk <- officer::fp_border(color = "#B5C3DF", style = "solid", width = 1)
BorderLt <- officer::fp_border(color = "#FFFFFF", style = "solid", width = 1)
flextable(arimapars) |> width(j=1:3, width=c(2,7,7), unit="cm") |> 
  padding(padding=2, part="all") |> border_outer(border = BorderDk, part = "all") |> 
  border_inner_v(border=BorderLt, part="all") |> set_header_labels(Info="Informed by") |> 
  set_caption(caption="Table 1: Definition of p, d, and q parameters in ARIMA time series forecasting models.", align_with_table=F, fp_p=officer::fp_par(text.align = "left", padding.bottom = 6))
```

&nbsp;

A periodic or *seasonal* ARIMA model (often called SARIMA) has a more
complex specification: <br> `ARIMA(p,d,q)(P,D,Q)(n)` , <br> where the
additional parameters refer to the seasonality: `P` is the number of
seasonal autoregressive predictors, `D` the seasonal differencing, `Q`
the seasonal moving averages, and `n` is the number of time steps per
period/season.

&nbsp;

### (a) With no seasonality

```{r simple arima, results='hold'}
am0 <- arima(x = soiltemp_T15_xts, order = c(1,0,3))
summary(am0)
confint(am0)
```

The output from `summary(am0)` shows the values and uncertainties of
the model parameters (`ar1, ma1, ma2, ma3, intercept`), the
goodness-of-fit parameters (we're interested in `aic`).

*Note*: `ar` parameters are auto-regression coefficients, and `ma` are 
moving average coefficients.

The output from `confint(am0)` is a table showing the 95% confidence
interval for the parameters. The confidence intervals should not
include zero! (if so, this would mean we can't be sure if the
parameter is useful or not).

&nbsp;

### (b) With seasonality

```{r seasonal arima, results='hold'}
ff <- findfrequency(soiltemp_T15_xts)
cat("Estimated time series frequency is",ff,"\n")
am1 <- arima(x = soiltemp_T15_xts, order = c(1,0,2),
             seasonal = list(order = c(1, 0, 1), period = ff))
summary(am1)
confint(am1)
```

*Note*: `sar` parameters are *seasonal* auto-regression coefficients, 
and `sma` are *seasonal* moving average coefficients.

The model which includes periodicity has the lowest AIC value (which
is not surprising, since the soil temperature data have clear
periodicity).

Checking residuals using the `checkresiduals()` function from the
`forecast` package is our best diagnostic tool to assess the validity
of our models. [This is a separate issue from how well the model
describes the data, measured by AIC. Also, we can't use `bptest()` to
analyse ARIMA residuals.]

In the output (plot and text) from `checkresiduals()`:
- the residual plot (top) should look like white noise
- the residuals should not be autocorrelated (bottom left plot)
- the p-value from the Ljung-Box test should be > 0.05 (text output)
- the residuals should be normally distributed (bottom right plot)

```{r diagnostic-arima, out.width="75%", fig.cap="Figure 14: Residual diagnostic plots for a non-seasonal ARIMA model of the  soil temperature time series.", results='hold'}
checkresiduals(am0)
```

&nbsp;

```{r diagnostic-sarima, out.width="75%", fig.cap="Figure 15: Residual diagnostic plots for a seasonal ARIMA (SARIMA) model of the  soil temperature time series.", results='hold'}
checkresiduals(am1)
```

&nbsp;

In both the diagnostic plots (Figure 14 and Figure 15) the residuals
appear to be random and normally distributed. There seems to be a more
obvious autocorrelation of residuals for the non-seasonal model,
reinforcing the conclusion from the AIC value that the **seasonal**
ARIMA model is the most appropriate.

&nbsp;

## Use the ARIMA model to produce a forecast using both models

**The whole point of generating ARIMA models** is so that we can 
attempt to forecast the future trajectory of our time series based on 
our existing time series data.

To do this, we use the appropriately named `forecast()` function from
the `forecast` package. The option `h =` is to specify how long we
want to forecast for after the end of our data &ndash; in this case
168 hours (same units as the periodicity from `findfrequency()`), that
is, one week.

<div style="border: 1px solid #039; background-color:#ffffe0; padding:
8px;"> <span style="font-size:12pt;">**One of the problems** we often
run into is the issue of **irregular time series**, where the
observations are not taken at regularly-spaced time intervals.
Features of time series, such as periodicity and the length of time we
want to forecast for, may be incorrect if our time series is
irregular.</span>

It is possible to "fill in" a time series so we have observations 
regularly spaced in time, using interpolation or numerical filtering 
methods. We won't be covering these methods here, though. 
</div>

&nbsp;

```{r forecast objects}
fc0 <- forecast(am0, h = 168)
fc1 <- forecast(am1, h = 168)
```

&nbsp;

### Then, look at forecasts with plots

```{r forecast-plots, fig.height=6, fig.width=7, out.width="65%", fig.align='center', fig.cap="Figure 16: Time series forecasts for soil temperature, based on (a) a non-seasonal ARIMA model and (b) a seasonal ARIMA model.", results='hold'}
par(mfrow = c(2, 1), cex.main = 0.9, mar = c(0,3,0,1), oma = c(3,0,1,0),
    mgp=c(1.6,0.3,0), tcl=0.2, font.lab=2)
plot(fc0,ylab = "Temperature (\u00B0C)",
     fcol = 7, xlab="", main = "", xaxt="n")
lines(soiltemp_T15_zoo, col = "grey70")
mtext("ARIMA with no seasonality", 1, -2.2, adj = 0.1, font = 2)
mtext(am0$call, 1, -1, adj = 0.1, col = 7, cex = 0.9, 
      family="mono", font = 2)
mtext("(a)", 3, -1.2, cex = 1.2, font = 2, adj = 0.95)

plot(fc1,ylab = "Temperature (\u00B0C)", main = "", fcol = 3)
lines(soiltemp$conc, col = "grey70")
mtext("Time since start (seconds)", 1, 1.7, cex = 1.2, font = 2)
mtext(paste("ARIMA with",am1$arma[5],"h periodicity"),
      side = 1, line = -2.2, adj = 0.1, font = 2)
mtext(am1$call, 1, -1, adj = 0.1, col = 3, cex = 0.9, family="mono")
mtext("(b)", 3, -1.2, cex = 1.2, font = 2, adj = 0.95)
```

&nbsp;

From the two plots in Figure 16, we observe that the seasonal ARIMA
model seems to reflect the form of the original data much better, by
simulating the diurnal fluctuations in soil temperature. Although the
forecasting uncertainty shown by the shaded area in Figure 16 is large
for both types of ARIMA model, the interval seems smaller for the
seasonal ARIMA model (Figure 16(b)).

```{r reset mfrow 2, message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1, 1))
```

We can also make a slightly 'prettier' time series forecast plot using the 
`autoplot()` function from the `ggplot2` package (Figure 17):

```{r forecast-ggplot, fig.height=4, fig.width=6, out.width='65%', fig.align='center', fig.cap="Figure 17: Time series forecast for soil temperature, based on  a seasonal ARIMA model, plotted using the ggplot2 `autoplot()` function.", results='hold'}
require(ggplot2) # gives best results using autoplot(...)
autoplot(fc1)+
  ylab("Temperature (\u00B0C)") +
  xlab(paste("Time since",index(soiltemp_T15_zoo)[1],"(s)")) +
  ggtitle("") +
  theme_bw()
```

&nbsp;
<div style="border: 2px solid #039; background-color:#ffc; padding: 8px;">
[ARIMA models are not the end of the time series modelling and forecasting story!]{style="font-family: serif; font-size: 14pt;"}
</div>

&nbsp;

## Exponential Smoothing Models

Sometimes, ARIMA models may not be the best option, and another
commonly used method is **exponential smoothing**.

Check `help(forecast::ets)` to correctly specify the model type using 
the `model =` option!

The easiest option (and the one used here) is to specify 
`model = "ZZZ"`, which chooses the type of model automatically.

```{r make ets model}
soiltemp_T15_ets <- ets(soiltemp_T15_xts, model ="ZZZ")
summary(soiltemp_T15_ets)
```

In this case the `ets()` function has automatically selected a model
(`ETS(A,Ad,N)`) with additive trend and errors, but no seasonality. We 
don't do it here, but we could manually select a model with 
seasonality included.

The default plot function in R can plot forecasts from exponential 
smoothing models too (Figure 18).

```{r ets-forecast, echo=-1, out.width="65%", fig.align='center', fig.cap="Figure 18: Forecast of soil temperature based on an automatically selected exponential smoothing model without a seasonal component."}
par(mfrow = c(1,1), mar = c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, oma = c(0,0,0,0), font.lab=2)
plot(forecast(soiltemp_T15_ets, h=168), col=8, fcol=6, 
     xlab = "Time since start (s)", ylab = "Temperature (\u00B0C)", main="")
mtext(names(soiltemp_T15_ets$par),3,seq(-1,-5,-1), adj = 0.7, col = 6, font = 3)
mtext(signif(soiltemp_T15_ets$par,3),3,seq(-1,-5,-1), adj = 0.8, col = 6, font = 3)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
par(mfrow = c(1,1), mar = c(4,4,1,1), oma = c(0,0,0,0))
```

&nbsp;

Exponential smoothing also [decomposes]{.underline} time series in a 
different way, into *level* and *slope* components (Figure 19).

```{r ets-decomposition, echo=-1, fig.height=6, fig.width=6, out.width='65%', fig.align='center', fig.cap="Figure 19: Exponential smoothing time series decomposition plots for soil temperature data."}
par(mfrow = c(1,1), mar = c(3,3,1,1), mgp=c(1.6,0.3,0), tcl=0.25, oma = c(0,0,0,0), font.lab=2)
plot(soiltemp_T15_ets, col=5, xlab="Time since start (s)") # ETS decomposition plots
```

&nbsp;

<p style="text-align: center; font-family: serif; font-size: 13pt;">(*Exponential smoothing is maybe not so good for the soil temperature data!*)</p>

<p id="References">&nbsp;</p>

# References

Australian Government (2023). [Assessment of change through monitoring data analysis](https://www.waterquality.gov.au/anz-guidelines/monitoring/data-analysis/data-assess-change){target="_blank"}, **In** [*Australian & New Zealand Guidelines for Fresh and Marine Water Quality*](https://www.waterquality.gov.au/anz-guidelines){target="_blank"}, [www.waterquality.gov.au/anz-guidelines/monitoring/data-analysis/data-assess-change](https://www.waterquality.gov.au/anz-guidelines/monitoring/data-analysis/data-assess-change){target="_blank"}.

Department of Water (2015). *Calculating trends in nutrient data*, Government of Western Australia, Perth.
[www.wa.gov.au/.../calculating-trends-nutrient-data](https://www.wa.gov.au/government/publications/calculating-trends-nutrient-data){target="_blank"}.

Hyndman R, Athanasopoulos G, Bergmeir C, Caceres G, Chhay L, 
O'Hara-Wild M, Petropoulos F, Razbash S, Wang E, Yasmeen F (2022). 
`forecast`:&nbsp;*Forecasting*
*functions for time series and linear models*. R package version 8.17.0,
[http://pkg.robjhyndman.com/forecast/](http://pkg.robjhyndman.com/forecast/){target="_blank"}.

Long, J.D., Teetor, P., 2019. Time series analysis. Chapter 14, 
*The R Cookbook*, Second Edition [https://rc2e.com/timeseriesanalysis](https://rc2e.com/timeseriesanalysis){target="_blank"}.

R Core Team (2022). `R`: *A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria. URL
[https://www.R-project.org/](https://www.R-project.org/){target="_blank"}.

Ryan, J.A. and Ulrich, J.M. (2020). `xts`: *eXtensible Time Series*. 
R package version 0.12.1.
[https://CRAN.R-project.org/package=xts](https://CRAN.R-project.org/package=xts){target="_blank"}

Zeileis, A. and Grothendieck, G. (2005). `zoo`: S3 Infrastructure for 
Regular and Irregular Time Series. *Journal of Statistical Software*, 
**14**(6), 1-27.
[doi:10.18637/jss.v014.i06](https://doi.org/10.18637/jss.v014.i06){target="_blank"}
