---
title: "Multivariate analysis"
subtitle: "Ordination analysis of species presence-absence data"
author: "Deirdre Gleeson and Andrew Rate"
date: "`r Sys.Date()`"
output: 
  html_document: 
    code_folding: show
    self_contained: no
    number_sections: no
    smart: no
    toc: no
---

## Learning Outcomes:

1. Students are able to appropriately analyse data from an eDNA field sampling campaign;
2. Students are able to interpret and understand statistical outputs from the R package `vegan`.

```{r set up, message=FALSE, warning=FALSE, results='hide'}
library(vegan)          # ecological multivariate analysis
library(flextable)      # publication-quality tables
  set_flextable_defaults(theme_fun = "theme_booktabs", font.size = 10)
library(officer)        # functions to support some flextable options
library(stringr)        # character string manipulation
library(viridis)        # colourblind-friendly colour palettes
library(scatterplot3d)  # used for 3D ordination plots
library(TeachingDemos)  # used for shadowtext() function
```

## Import the data

**Notes**:

1. This code imports the data from a file where the **samples are rows** and the **species are columns**.
2. We convert the column `Site` to a *factor* (categorical information in R)

```{r read-curate-data, message=FALSE, warning=FALSE, results='hide'}
git <- "https://github.com/Ratey-AtUWA/Learn-R-web/raw/main/"
eDNA2022 <- read.csv(file = paste0(git,"eDNA2022.csv"))
eDNA2022$Site <- as.factor(eDNA2022$Site)
```

If you have *species* as rows and *samples* as columns, we will need to do a bit
of wrangling!

```{r read and wrangle species as rows data, eval=FALSE}
speciesAsRows <- read.csv(file = paste0(git,"speciesAsRows.csv"), row.names = 1)
speciesAsRows <- rbind(str_remove(colnames(speciesAsRows), "X"),
                       speciesAsRows)
row.names(speciesAsRows)[1] <- "SiteID"
speciesAsRows <- as.matrix(speciesAsRows) # convert to matrix for next step
speciesAsRows <- t(speciesAsRows) # t() means transpose
row.names(speciesAsRows) <- NULL
eDNA2022 <- as.data.frame(speciesAsRows)
eDNA2022$Site <- as.factor(eDNA2022$Site)
```

## List of all the species and their common names

```{r tabulate-species, message=FALSE, warning=FALSE, results='hold'}
altnames <- 
  c("MosqFish", "Goby", "Mullet", "Bream", "BnjoFrog", "GToadlet", "MlrdDuck",
    "SpotDove", "SwampHen", "Crake", "WFHeron", "Lorikeet", "Cattle", "Sheep",
    "Mouse", "Rat", "HydroidB", "HydroidO", "BwnJelly", "MoonJely", "Dog")
names_table <- 
  data.frame(Taxonomic = gsub("."," ",colnames(eDNA2022[,6:26]), fixed = T),
             Common=c("Mosquito Fish","Bridled Goby","Gray Mullet","Black Bream",
                      "Western Banjo Frog", "Gunther's Toadlet", "Mallard Duck",
                      "Spotted Turtle Dove", "Purple Swamphen", "Spotted Crake",
                      "White Faced Heron", "Rainbow Lorikeet", "Cattle", "Sheep",
                      "Mouse", "Rat", "HydroidB", "HydroidO", "Brown Jellyfish",
                      "Moon Jellyfish", "Domestic Dog"),
             Abbreviated=altnames)
flextable(names_table,cwidth = c(3,2,2)) |> italic(j=1, italic=T) |> bold(bold=T, part="header") |> 
  padding(padding=2, part="all") |> 
  set_caption(caption="Table 1: Taxonomic, common, and abbreviated names for the species detected by eDNA sampling and analysis at Ashfield Flats in 2022.", align_with_table = F, fp_p=fp_par(text.align = "left", padding.bottom = 6))
```

&nbsp;

We need to make a subset of our imported data which contains just the species
presence/absence information (not sample IDs, sites, reps). We also overwrite
the original column names with abbreviated names for better visualisation later.

```{r generate all species data, results='hold'}
allSpecies <- eDNA2022[,6:26] # not columns 1:5 which are sample IDs etc.
colnames(allSpecies) <- names_table[,3]
```

## Difference and Similarity measures

We start by assessing how similar or different our samples are – to do this we
need to set up a resemblance matrix. Ideally our 4 replicated samples are more
similar to one another than any other samples – but with field sampling this is
not always the case. To determine similarity we need to use a similarity
measure.

### Similarity Measures:
A similarity measure is conventionally defined to take values between 0 and 100%
(or 0 and 1) with the ends of the range representing the extreme possibilities:

- Similarity = 100% (*or 1*) if two samples are totally similar
- Similarity = 0 if two samples are totally dissimilar

What constitutes total similarity, and particularly total dissimilarity, of two
samples depends on the specific criteria that are adopted. For example most
biologists would feel that similarity should equal zero when two samples have no
species in common, and similarity should equal 100% if two samples have
identical entries. Similarities are calculated between every pair of samples
creating a similarity matrix. There are a number of similarity measures (ways to
calculate similarity) which can be used to create this matrix.

**Bray Curtis** and **Jaccard** are two different similarity measures we can use
-– both are commonly used in ecology and have desirable properties &ndash; similarity
is calculated based on the taxa that samples have in common. Similarity takes a
value of 0 if two samples have no species in common and joint absences have no
effect (*i.e*. a taxon jointly absent in two samples does not contribute to its
similarity). The difference between them is that Bray Curtis will use abundance
information (where it is present) in assessing similarity, whereas Jaccard
converts abundances to binary (*i.e*. 0s and 1s, with 1 representing presence
and 0 meaning absence) to assess similarity. In our case it does not matter much
which we use, since we have already converted to binary – so let’s proceed with
Jaccard.

**Ecological Question – If two samples have no taxa present each does this make them more similar to one another?**

To assess similarity in the `vegan` **R** package carry out the following steps. The
output shows the first few rows and columns of the dissimilarity matrix; each
number represents the Jaccard distance between samples based on the presence or
absence of species.

```{r dissim matrix all Jaccard, eval=FALSE}
AF_diss_all <- vegdist(allSpecies, 
                           distance = "jaccard") # default is distance = "bray"

# just look at the top corner of the dissimilarity matrix:
txt0 <- capture.output(print(AF_diss_all, digits=3)); txt0[1:10]; rm(txt0)
```

```{r dist matrix all jaccard output with cat, echo=FALSE}
AF_diss_all <- vegdist(allSpecies, 
                           distance = "jaccard")
txt0 <- capture.output(print(AF_diss_all,digits=3))
for (i in 1:11){ cat(txt0[i],"\n") }
```

# Ordinations

Next we will construct our ordination – a procedure to visualise the similarity
matrix by simplifying (scaling) it into fewer dimensions than our original data.
This reduction of dimensions is useful as the original data will have as many
dimensions as there are species, which is impossible to visualise! There are two
different types of ordination that we will use: **Multi-Dimensional-Scaling**
(nmMDS) and **Principal Coordinates Analysis** (PCO). If you have previously
used PCA &ndash; Principal Component Analysis &ndash; you may be familiar with
ordination methods. PCA is useful for environmental datasets &ndash; these are
represented by a similarity matrix derived via use of Euclidean Distance as a
measure of similarity. PCA is more suited to environmental variables because of
the type of data – there are no large blocks of zeros (usually) and it is no
longer necessary to select a similarity measure that ignores joint absences.
When looking at ordination plots in general you should note that sample points
that are spatially located together share common characteristics and sample
points that are distantly located from one another share less characteristics.
This holds whether we are working with environmental variables (pH, EC, moisture
content, heavy metals *etc*.) or species abundances.

The main difference between MDS and PCO relates to the underlying use of the
resemblance matrix.

**MDS** takes the similarity measures calculated by Jaccard or Bray Curtis and
ranks them in order. Thus pairs are samples are ranked according to highest to
lowest similarity and are thus plotted in multidimensional space. Since it uses
the rank and not the actual values, it's commonly called "non-[para]metric"
multi-dimensional scaling, or **nmMDS**.

**PCO** takes the actual values of the underlying measure of similarity and
plots pairs of samples in multi-dimensional space. It will then look for the
direction (vector) of greatest variance in the multi-dimensional "cloud" of
points, which becomes the first PCO dimension. Successive dimensions explain the
most possible remaining variance, until all variance is accounted for. So, many
PCO dimensions are possible, but usually only the first 2 or 3 are useful for
visualizing our data.

For both types of ordinations the software attempts to preserve the similarity
of each pair of samples &ndash; while this is feasible in multidimensional
space, in order for us to view the ordination we see this in 2 or 3 dimensions
&ndash; thus the software attempts to preserve as much as is feasible of the
sample pair similarity as calculated &ndash; but in reality we need to consider
how much the 2- dimensional view is a true representation of the data set. In
nmMDS we use the *stress* value to evaluate this, in PCO we can plot multiple
axes, *i.e*. axis 1 *vs*. axis 2, axis 2 *vs*. axis 3 and so on. The amount of 
variance explained in each PCO dimension is called the *eigenvalue*, and the
output will include these values; the greater the eigenvalue, the more useful a
PCO dimension is. In practice, much of the variability is often explained in the
first two axes (*i.e*. 2 dimensions) and we will limit our analysis to these
axes for this dataset.

## Non-metric multidimensional scaling (nmMDS)

In vegan this is done using the `metaMDS()` function. We need to specify an
input matrix of just the species data (*e.g*. `allSpecies` which we made
earlier). The default is to find 2 dimensions (`k`) using a Bray-Curtis
dissimilarity matrix, but we can set these options to different values using 
`k =` and `distance =` options. The dissimilarity matrix is
calculated within the `metaMDS()` function, so for consistency we set 
`distance = "jaccard"`.

```{r run nmds all jaccard, results='hide'}
# default with 2 dimensions
AF_nmds_all <- metaMDS(allSpecies, distance = "jaccard")
# 3 dimensions for comparison
AF_nmds_all3 <- metaMDS(allSpecies, k=3, distance = "jaccard")
```

```{r show mds all jaccard}

# show output for 2 nmMDS dimensions
AF_nmds_all
```

The 2D stress value is &sime;&nbsp;0.16. This can be interpreted as follows:

```{r stress-range-diagram, fig.height=1., fig.width=6, out.width="50%", fig.align='center', echo=FALSE}
par(mar=c(2,0,1,0))
plot(c(0,0.4),c(0,1),type="n",bty="n",xaxs="i",yaxs="i",xaxt="n",yaxt="n",xlab="",ylab="")
axis(1,at=c(0.05,0.1,0.2,0.3),labels=c(0.05,0.1,0.2,0.3));abline(h=0)
mtext("What's good stress? (for nmMDS)", adj=0.02)
abline(v=c(0.05,0.1,0.2), lty = 2)
text(c(0.025,0.075,0.15,0.3), rep(0.25,5),
     labels=c("excellent","great","good but caution",
              "maybe useful but potentially wrong . . ."),
     col = c("chartreuse4","blue","darkorange2","firebrick3"))
par(mar=c(4,4,1,1))
```

The stress value is relatively high as we are dealing with a large number of
samples with lots of values of `0` or `1` in a 2 dimensional space – resulting in a
challenging similarity environment.

$\Rightarrow$ Check the stress value of the 3-D ordination – is it lower? Remember that we made this object in the code block above:

```{r show mds all jaccard 3dim, results='hide'}
AF_nmds_all3
```

Figure 1 below mimics the nmmds plot in Deirdre's original `PRIMER` handout
(note that the sign of dimension MDS2 is reversed)

```{r plot-nmds-all-Jaccard-2D, fig.height=6, fig.width=6, fig.cap="Figure 1: 2D NMDS ordination plot based on Jaccard (dis)similarities in eDNA species presence-absence data for water sampled from Ashfield Flats in 2022.", out.width="60%"}
palette(c("black", viridis(10))) # a viridis palette
par(mfrow=c(1,1), mar=c(3,3,2,0.5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
plot(AF_nmds_all$points, pch=c(rep(21:25, 2),21)[eDNA2022$Site], 
     xlim = c(-1.6,2.2), ylim = c(1.2,-1.2), 
     bg=seq(1,11)[eDNA2022$Site], 
     cex = 1.4, main = "All species", col.main = "steelblue")
text(AF_nmds_all, display = "species", col = "#a0006080", font=3, cex = 1)
legend("bottomright", box.col="grey", bg="transparent", ncol=4, title = "Site", 
       legend=seq(1,11), pch=c(rep(21:25, 2),21), 
       inset=0.01, pt.bg=seq(1,11), pt.cex=1.4, cex=1.2)
```

&nbsp;

You can play around with the code for Figure 1 to make additional changes – you
won’t break it by trying, so play around as much as you like. You will notice
that there are only 2 points for Site 2 – this is because the points are
overlaid on top of one another – as we are picking up only a few taxa some of
the samples are very similar to one another. However we can generally see that
there are sample groupings and that likely location has an effect (i.e. samples
that are from the same location are closer together). We will test this
statistically a little later.

We can also generate a form of 3-dimensional plot using the `scatterplot3d()` 
function from the `scatterplot3d` R package. Figure 2 shows a 3D plot for the 
3-dimension NMDS generated above.

```{r plot-nmds-3D, fig.height=5, fig.width=6.5, out.width="70%", fig.cap="Figure 2: 2D NMDS ordination plot based on Jaccard (dis)similarities in eDNA species presence-absence data for water sampled from Ashfield Flats in 2022."}
palette(c("black", viridis(10))) # a viridis palette
par(mfrow=c(1,1), mar=c(3,3,2,0.5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
s3d <- scatterplot3d(x=AF_nmds_all3$points[,1], y=AF_nmds_all3$points[,2],
              z=AF_nmds_all3$points[,3],
              xlab="MDS1", ylab="MDS2", zlab="MDS3",
              angle=25, mar=c(4,4,1,2), type="h", col.axis="darkgray", color="gray",
              pch=c(rep(21:25, 2),21)[eDNA2022$Site],
              bg=c(1:11)[eDNA2022$Site], cex.symbols=1.4)
legend(s3d$xyz.convert(1.7, 0.75, 1.6), legend = levels(eDNA2022$Site), title="Site", 
       bg="#ffffffa0", box.col="transparent", 
       pt.bg= c(1:11), pt.cex=1.4, col="gray", pch=c(rep(21:25, 2),21))
```

## Principal Coordinates (PCoA) Analysis

In the 'vegan' package we use Weighted classical multidimensional scaling, also
known as *Weighted Principal Coordinates Analysis*. This is implemented using 
the function `wcmdscale()`. We plot the PCO output in Figure 3 below.

```{r run-pco, message=FALSE, warning=FALSE, results='hold'}
AF_pco_all <- wcmdscale(vegdist(allSpecies), eig=TRUE)
AF_pco_all
```

```{r basic-pco-plot, fig.height=4, fig.width=4, fig.cap="Figure 3: Basic principal coordinates analysis (PCoA) ordination plot based on Jaccard (dis)similarities in eDNA species presence-absence data for water sampled from Ashfield Flats in 2022."}
par(mfrow=c(1,1), mar=c(3,3,2,0.5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
plot(AF_pco_all)
```

Figure 3 is a basic PCoA plot based on all species, showing the locations of
individual samples by row number.

We can make more informative plots by using the information stored in the
output object (`AF_pco_all`) &ndash; see Figure 4.

```{r better-pco-plot-s3d, fig.width=4.5, fig.height=4, out.width="70%", fig.cap="Figure 4: Principal coordinates (PCoA) ordination plot in 3D based on Jaccard (dis)similarities in eDNA species presence-absence data for water sampled from Ashfield Flats in 2022.", echo=1:4, out.width="50%"}
palette(c("black", plasma(10))) # a viridis palette
par(mfrow = c(1,1), mar = c(3,3,1,1), mgp = c(1.5,0.2,0), tcl = 0.25, 
    font.lab = 2, xpd=T)
s3d <- scatterplot3d(x=AF_pco_all$points[,1], y=AF_pco_all$points[,2], z=AF_pco_all$points[,3],
              xlab="Dim 1", ylab="Dim 2", zlab="Dim 3",
              angle=65, mar=c(4,4,3,2), type="h", col.axis="gray", color="#e0e0e0",
              pch=c(rep(21:25, 2),21)[eDNA2022$Site],
              bg=c(1:11)[eDNA2022$Site], cex.symbols=1.4,
              xlim=c(-0.6,0.4),ylim=c(-0.6,0.4),zlim=c(-0.3,0.4))
legend(s3d$xyz.convert(-1.5, 0.4, 0.7), legend = levels(eDNA2022$Site), title="Site", 
       horiz=TRUE, cex=0.8, bg="#ffffffa0", box.col="transparent", 
       pt.bg= c(1:11), pt.cex=1.1, col="gray", pch=c(rep(21:25, 2),21))
# plot(AF_pco_all$points[,c(1,2)], xlim = c(-0.6,0.4), ylim = c(-0.6,0.4), 
#      pch=19, cex = 1, col = 4)
# text(AF_pco_all$points[,c(1,2)],labels = eDNA2022$SiteID,
#      col=seq(1,11)[eDNA2022$Site])
# mtext("(a)", line=-1.3, adj=0.05, cex=1.2)
# plot(AF_pco_all$points[,c(1,3)], xlim = c(-0.6,0.4), ylim = c(-0.6,0.4), 
#      pch=15, cex = 1, col = 6)
# text(AF_pco_all$points[,c(1,3)],labels = eDNA2022$SiteID,
#      col=seq(1,11)[eDNA2022$Site])
# mtext("(b)", line=-1.3, adj=0.05, cex=1.2)
par(xpd=F)
```

```{r calculate pco scores but not sure if we need this}
AF_pco_scores <- scores(AF_pco_all)
```

<hr />

# PERMANOVA

In order to test where location has a significant effect on the taxa at our
field site we need to run an ANOVA – however as we have multiple species to test
together we cannot run a basic ANOVA – we must run a multivariate permutational
ANOVA or **PERMANOVA**.

PERMANOVA shares some resemblance to ANOVA where they both measure the
sum-of-squares within and between groups, and make use of an F test to compare
within-group to between-group variance. However, while ANOVA bases the
significance of the result on assumption of normality, PERMANOVA draws tests for
significance by comparing the actual F test result to that gained from random
permutations of the objects between the groups. Moreover, whilst PERMANOVA tests
for similarity based on a chosen distance measure, ANOVA tests for similarity of
the group averages.

We could calculate a univariate measure of diversity (a Simpson or Shannon
index, for example) and run an ANOVA on that univariate value. In doing so,
however, we are reducing the complex dataset to a simplified version and losing
discrete information in the process. Remember that ANOVA also has an underlying
assumption of normality &ndash; generally speaking, biological species data are
not normally distributed and thus do not satisfy this assumption. This makes
applying a permutational ANOVA a much better option. Before the advent of
computational capacity a permutational ANOVA would have taken quite some time to
run &ndash; luckily now it only takes a matter of seconds to minutes depending
on the complexity of the dataset.

In the `vegan` package, the `adonis2()` function implements Permutational
Multivariate Analysis of Variance Using Distance Matrices (*i.e*. PERMANOVA).

We use a formula in the `adonis2()` function which expresses the community matrix 
(*i.e*. the columns of species presence or absence) as a function of a factor 
(in this case, `Site`).

```{r permanova-main-effect}
AF_permanova_all <- adonis2(eDNA2022[,6:26] ~ Site, data = eDNA2022,
                            permutations = 9999)
AF_permanova_all
```

P value by permutation =`0.0001` indicating a significant effect of location on
the community composition of our samples. Note that 9999 permutations were
undertaken to arrive at this value.

The output from the `pairwise.adonis2()` function is quite lengthy (to see it, 
run `print(AF_PWpermanova_all)`). We can tidy it up for clarity with a custom 
function (`plainPW2()` &ndash; see output below).

```{r pairwise-Permanova, echo=2:6, message=FALSE, warning=FALSE, paged.print=FALSE}
options(width = 120)
source("https://github.com/Ratey-AtUWA/eDNA/raw/master/FUN_pairwise_adonis2.R")
source("https://github.com/Ratey-AtUWA/Learn-R/raw/main/plain-pairwise-FUNCTION2.R")
AF_PWpermanova_all <- pairwise.adonis2(eDNA2022[,6:26] ~ Site, data = eDNA2022)
AF_plainPW_all <- plainPW2(AF_PWpermanova_all, ns.repl=TRUE, alpha=0.05) # see custom function code below
s0 <- data.frame(Site=1:10)
flextable(cbind(s0,AF_plainPW_all)) |> width(width=0.5) |> theme_booktabs() |> 
  bold(bold=T,part="header") |> bold(bold=T, j=1) |> 
  padding(padding=2, part="all") |> align_text_col(align="right") |> 
  set_caption(caption="Table 2: Pairwise comparison matrix based on PERMANOVA testing the effect of `Site` on multivariate species presence-absence (ns = p > 0.05).", align_with_table = F, fp_p=fp_par(text.align = "left", padding.bottom = 6))
```

<p>&nbsp;</p>

**Notes**: 

- the pairwise Adonis2 source code is by Pedro Martinez Arbizu at [https://github.com/pmartinezarbizu/pairwiseAdonis](https://github.com/pmartinezarbizu/pairwiseAdonis){target="_blank"}.
- The code for the `plainPW2()` function is as follows:

```{r plainPW2-function-code, eval=FALSE, paged.print=FALSE}
# function inputs output object from Pedro Martinez Arbizu's pairwise.adonis2()
# code at https://github.com/pmartinezarbizu/pairwiseAdonis
plainPW2 <- function(x, ns.repl=TRUE, alpha=0.05){
  tablout <- data.frame(Pair=rep(NA, length(x)-1),
                        P_value=rep(NA, length(x)-1))
  for(i in 2:length(x)){
    tablout[i-1,] <- c(names(x)[i],
                       as.data.frame(x[names(x)[i]])[1,5])
    }
  n0 <- (ceiling(sqrt(length(x)*2)))-1
  ptable <- as.data.frame(matrix(rep(NA, n0^2), ncol = n0))
  colnames(ptable) <- c(1:n0)+1
  r0 <- 1 ; rn <- n0
  for(i in 1:n0){
    ptable[i,] <- c(rep(NA,n0-((rn-r0)+1)),as.numeric(c(tablout[r0:rn,2])))
    r0 <- rn+1
    rn <- rn+(n0-i)
  }
  if(ns.repl==TRUE){
    for(i in 1:ncol(ptable)){
      ptable[which(ptable[,i]>alpha),i] <- "ns"
      }
    }
  return(ptable)
}
```

You can also get the `plainPW2()` function code at
[github.com/Ratey-AtUWA/.../plain-pairwise-FUNCTION2.R](https://raw.githubusercontent.com/Ratey-AtUWA/Learn-R/main/plain-pairwise-FUNCTION2.R){target="_blank"}.

&nbsp;

# Extras

# Additional options for `vegan` ordination plots

If we have factors (categories) in our data it's useful to show these using our
multivariate ordination plots. We can make a rough categorisation of Sites based
on their site location, for example as shown in the code below:

```{r make-N-S-factor, message=FALSE, warning=FALSE, results='hold'}
eDNA2022$Zone <- cut(eDNA2022$Northing, breaks = c(1,6468100,9999999), 
                     labels=c("South","North"))
(eDNA2022$Zone <- factor(eDNA2022$Zone, levels=c("North","South"))) # re-order
```

We can then add geometric shapes on our ordination plots to emphasise the
positions of different groups (*i.e*. factor levels) within the ordination
dimensions' space (Figure 5). These plot features are really only applicable to
two-dimensional (*x-y*) plots &ndash; 3D geometric shapes could be hard to
interpret on a 3D plot!.

```{r ellipses-hulls-spiders, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, fig.cap="Figure 5: Ordination plots made using a combination of functions from the `vegan` package and base-R, showing (a) 75% CI ellipses around Zone groups; (b) convex hulls around Site groups; (c) ‘spiders’ around Zone groups. The `shadowtext()` function from the `teachingDemos::` package is used for clearer labels. The Zones in this Figure are just for illustration and based on whether the site is south or north of an arbitrary coordinate, and may make no sense!", results='hold'}
# plot (a) ellipses
par(mar=c(3,3,1,1), mgp=c(1.7,0.3,0), tcl=0.25, lend=2, ljoin=1, font.lab=2,
    mfrow=c(2,2))
palette(c("black", inferno(6)[2:6],"white"))
plot(AF_nmds_all, display="sites", cex=1.4, xlim=c(-1.35,1.65), ylim=c(-1.5,1.5))
mtext("(a)", side=3, line=-1.6, adj=0.03, cex=1.4)
points(AF_nmds_all$points, col=c(2,5)[eDNA2022$Zone], pch=19, cex=1.4)
ordiellipse(AF_nmds_all, groups=eDNA2022$Zone, col=c(2,5), lwd=2, 
            kind = "sd", conf=0.75)
shadowtext(tapply(AF_nmds_all$points[,1], eDNA2022$Zone, mean),
           tapply(AF_nmds_all$points[,2], eDNA2022$Zone, mean),
           labels=levels(eDNA2022$Zone), cex=1.2, col=c(2,5), bg=7, r=0.2)
legend("bottomright", bty="n", legend=levels(eDNA2022$Zone), title="Zone",
       col=c(2,5), pt.bg=c(2,4,5), lty=1, lwd=2, pch=NA, pt.cex=1.5, cex=1.2)
legend("bottomright", bty="n", legend=levels(eDNA2022$Zone), col=7, 
       pt.bg=c(2,5), lty=NA, lwd=2, pch=21, pt.cex=2.2, cex=1.2)

# plot (b) convex hulls
# par(mar=c(3,3,1,1), mgp=c(1.7,0.3,0), tcl=0.25, lend=2, ljoin=1, font.lab=2)
palette(c("black", magma(5)[2:5],"white"))
plot(AF_nmds_all, display="sites", cex=1.4, xlim=c(-1.35,1.65), ylim=c(-1.5,1.5))
ordihull(AF_nmds_all, groups=eDNA2022$Zone, 
         col=c(2,4), lwd=2, label=F)
points(AF_nmds_all$points, bg=c(2,4)[eDNA2022$Zone], 
       pch=c(21,22)[eDNA2022$Zone], cex=2, col=6)
shadowtext(tapply(AF_nmds_all$points[,1], eDNA2022$Zone, mean),
           tapply(AF_nmds_all$points[,2], eDNA2022$Zone, mean),
           labels=levels(eDNA2022$Zone), cex=1.2, col=c(2,4), bg=6, r=0.2)
mtext("(b)", side=3, line=-1.6, adj=0.03, cex=1.4)
legend("bottomright", bty="n", legend=levels(eDNA2022$Zone), title="Zone",
       col=c(2,4), pt.bg=c(2,4), lty=1, lwd=2, pch=NA, pt.cex=1.5, cex=1.2)
legend("bottomright", bty="n", legend=levels(eDNA2022$Zone), col=6, 
       pt.bg=c(2,4), lty=NA, lwd=2, pch=c(21,22), pt.cex=2.2, cex=1.2)

# (c) spiders
palette(c("black", plasma(6)[2:6],"white"))
plot(AF_nmds_all, display="sites", cex=0.8, xlim=c(-1.35,1.65), ylim=c(-1.5,1.5))
mtext("(c)", side=3, line=-1.6, adj=0.03, cex=1.4)
points(AF_nmds_all$points, col=c(2,4)[eDNA2022$Zone], pch=19, cex=1.4)
ordispider(AF_nmds_all, groups=eDNA2022$Zone, col=c(2,4), lwd=1)
shadowtext(tapply(AF_nmds_all$points[,1], eDNA2022$Zone, mean),
           tapply(AF_nmds_all$points[,2], eDNA2022$Zone, mean),
           labels=levels(eDNA2022$Zone), cex=1.2, col=c(2,4), bg=7, r=0.2)
legend("bottomright", bty="n", legend=levels(eDNA2022$Zone), title="Zone",
       col=c(2,4), pt.bg=c(1,3,5), lty=1, lwd=2, pch=NA, pt.cex=1.5, cex=1.2)
legend("bottomright", bty="n", legend=levels(eDNA2022$Zone), col=7, 
       pt.bg=c(2,4), lty=NA, lwd=2, pch=21, pt.cex=2.2, cex=1.2)

plot(c(0,1),c(0,1),bty="n",axes=F,ann=F,type="n")
text(0.5,0.5, cex=1.5, col=3, font=3,
       labels = "The categories\n'North' and 'South' may\nnot be useful zones!")
```

&nbsp;

### Stress plot for nmMDS based on Jaccard dissimilarities, all species

```{r stress-plot, fig.height=5, fig.width=7, fig.align='center', out.width="60%", fig.cap="Figure 6: Stress plot showing the goodness-of-fit between actual dsiimilarity for the Ashfield Flats 2022 eDNA data and dissimilarities predicted by NMDS ordination distances.", message=FALSE, warning=FALSE}
par(mfrow=c(1,1), mar=c(3,3,0.5,0.5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
stressplot(AF_nmds_all)
```

## Analysis with just the aquatic species

These are: *Gambusia holbrooki, Arenigobius bifrenatus, Mugil cephalus, Acanthopagrus butcheri, Limnodynastes dorsalis, Pseudophryne guentheri, Blackfordia polytentaculata, Obelia bidentata, Phyllorhiza punctata, Aurelia sp.*

```{r subset to just aquatic, paged.print=FALSE, results='hold'}
fish <- allSpecies[,c(1:6,17:20)]
head(fish)
```

### for interest look at the dissimilarity matrix

```{r dissim matrix aquat bray, eval=FALSE}
AF_diss_aquat <- vegdist(fish, 
                           distance = "bray")
txt0 <- capture.output(print(AF_diss_aquat,digits=3));txt0[1:10];rm(txt0)
```

```{r dist matrix aquat bray output with cat, echo=FALSE}
AF_diss_aquat <- vegdist(fish, 
                           distance = "bray")
txt0 <- capture.output(print(AF_diss_aquat,digits=3))
for (i in 1:11){ cat(txt0[i],"\n") }
```

```{r run nmds aquat Bray, results='hide'}
AF_nmds_aquat <- metaMDS(fish, trymax = 500, 
                           distance = "jaccard")
```
```{r show nmds aquat Bray}
AF_nmds_aquat
```
```{r plot nmds aquat Bray, fig.height=4, fig.width=4, fig.align='center', fig.cap="Figure 7: Basic NMDS plot showing observations as symbols and species as text. Presence-absence data from eDNA sampling and analysis at Ashfield Flats."}
par(mfrow=c(1,1), mar=c(3,3,2,0.5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
plot(AF_nmds_aquat, type="p",display = "sites", cex = 1.2)
text(AF_nmds_aquat, display = "species", col = "royalblue", cex = 0.8)
```

### alternative to vegan plots
```{r alt-to-vegan-nmds-aquat, fig.height=3.5, fig.width=5, fig.cap="Figure 8: Two-dimensional non-metric multidimensional scaling (NMDS) plot showing observations as symbols and species as text located in MDS2 vs. MDS1 space. Presence-absence data for aquatic species only from eDNA sampling and analysis at Ashfield Flats.", echo=1:5}
palette(c("black", plasma(10))) # a viridis palette
par(mfrow=c(1,1), mar=c(3,3,0.5,5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2, xpd=T)
plot(AF_nmds_aquat$points, pch=c(rep(21:25, 2),21)[eDNA2022$Site], 
     xlim = c(-1.2,1.8), 
     bg = seq(1,11)[eDNA2022$Site], 
     cex = 1.2, col.main = "steelblue")
text(AF_nmds_aquat, display = "species", col = "grey33", font=3, cex = 0.8)
legend(2.2,1.2, legend=seq(1,11), bty="n", title="Site", cex=1, pt.cex=1.2, 
       pch=c(rep(21:25, 2),21), pt.bg=1:11)
par(xpd=F)
```

## Detrended correspondence analysis - all species

```{r run-DCA, message=FALSE, warning=FALSE, results='hold'}
AF_dca_all <- decorana(allSpecies)
AF_dca_all
```

```{r decorana-plot, fig.height=4, fig.width=4, fig.cap="Figure 9: Two-dimensional detrended correspondence analysis (DCA) plot showing observations as symbols and species as text. Presence-absence data from water sampling and eDNA analysis at Ashfield Flats."}
par(mfrow=c(1,1), mar=c(3,3,0.5,0.5), mgp=c(1.5,0.2,0), tcl=0.2, font.lab=2)
plot(AF_dca_all, col=c(1,4))
```

## For comparison if wanted: nmMDS again with Bray-Curtis distances
```{r run nmds Bray allspecies, results='hide'}
AF_nmds_all_B <- metaMDS(allSpecies) # uses Bray by default
```

```{r show nmds Bray allspecies, results='hold'}
AF_nmds_all_B
```

The results should be identical to nmMDS with Jaccard, since the community
matrix is already converted to presence-absence (binary ones and zeros).

&nbsp;

# References and `R` Packages

Dunnington, Dewey (2017). `prettymapr`: *Scale Bar, North Arrow, and Pretty Margins in R*. R package version 0.2.2.
[CRAN.R-project.org/package=prettymapr](https://cran.r-project.org/package=prettymapr){target="_blank"}.

Giraud T (2021). `maptiles`: *Download and Display Map Tiles*. R package version 0.3.0, [CRAN.R-project.org/package=maptiles](https://cran.r-project.org/package=maptiles){target="_blank"}.

Garnier S, Ross N, Rudis R, Camargo AP, Sciaini M, Scherer C (2021). *Rvision - Colorblind-Friendly Color Maps for R* (`viridis`). R package version 0.6.2. [sjmgarnier.github.io/viridis/](https://sjmgarnier.github.io/viridis/){target="_blank"}

Gohel D, Skintzos P (2022). `flextable`: *Functions for Tabular Reporting*. R package version 0.8.1, [CRAN.R-project.org/package=flextable](https://cran.r-project.org/package=flextable){target="_blank"}.

Oksanen J, Simpson G, Blanchet F, Kindt R, Legendre P, Minchin P, O’Hara R, Solymos P, Stevens M, Szoecs E, Wagner H, Barbour M, Bedward M, Bolker B, Borcard D, Carvalho G, Chirico M, De Caceres M, Durand S, Evangelista H, FitzJohn R, Friendly M, Furneaux B, Hannigan G, Hill M, Lahti L, McGlinn D, Ouellette M, Ribeiro Cunha E, Smith T, Stier A, Ter Braak C, Weedon J (2022). `vegan`: *Community Ecology Package*. R package version 2.6-2, [CRAN.R-project.org/package=vegan](https://cran.r-project.org/package=vegan){target="_blank"}.

Pebesma, E., 2018. Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal* **10**(1), 439-446, [doi:10.32614/RJ-2018-009](https://doi.org/10.32614/RJ-2018-009){target="_blank"}. (package `sf`)

Wickham H (2022). `stringr`: *Simple, Consistent Wrappers for Common String Operations*. R package version 1.4.1, [CRAN.R-project.org/package=stringr](https://cran.r-project.org/package=stringr){target="_blank"}.

