---
title: "Workshops 1 and 2"
author: "Andrew Rate"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

**Scope from LMS Class Schedule**

#### Workshop 1

* Getting started with R Projects and R Markdown
* R operators, classes
* R data structures
* Input/output

#### Workshop 2
* row/column names
* indices - locating values inside R objects
* coercions - forcing an R object to be another class
* summaries of data objects

## R Classes and data structures

R works with *objects*, which are named pieces of information held in
your computer's memory or *workspace*. Objects can hold different
*classes* of information, such as numbers, text, or logical
(TRUE/FALSE) values. The code chunks below let you practice doing some
of this.

Note that any line of code beginning with `#` is a *comment* in R, and
does not do anything except provide information to the reader.

We put any code (and comments) in _code chunks_ which begin with
‵‵‵{r} and end with ‵‵‵ (sometimes additional options can go after the
`r` in `{r}`). Code chunks are in shaded boxes in the R markdown file.
The code chunks can be run (click ▶ ), 
and any output will appear in the unshaded box below each code chunk.

```{r}
# make an object x with a single number in it

# the symbol `<-` means 'is assigned the value' or 'becomes'

# so x <- 12 means "the object x is assigned the value 12"

x <- 12
x
class(x)
```

Some notes here:

* if we type the name of an object, we display its contents.
  Running the name `x` as R code is a shortcut for `print(x)`.
  
* `class(x)` is our first use of an R *function*.
  Functions do something for us when we run them -- in this case, the
  function `class()` produces output telling us what class of object 
  we have. `class` is the function name, and it is always followed by
  parentheses `()` which contain the *argument(s)* or input to the
  function.

```{r}
# make an object y with several numbers in it
# the simple function c() means "combine"
y <- c(12, 31, 8, 51, 59, 23)
y
class(y)
```

```{r}
# make an object z with several strings (text values) in it
z <- c("Monday", "Tuesday", "Hydrogen", "Helium", "Mercury", "Venus")
z
class(z)
```

```{r}
# make an object q with several logical (TRUE/FALSE) values in it
q <- c(TRUE, FALSE, T, F)
# we can use T and F as abbreviated forms of logical values
q
class(q)
```

```{r}
# make an object k with a mix of numbers, logical, and strings
k <- c(1, 2, TRUE, FALSE, "Mercury", "Venus")
k
class(k)
```

This shows that the object that we create with the `c()` function can
not store different types of values, just a single class.

The objects we have created are *vectors*, or one-dimensional arrays
of values belonging to only one class.

Later, in the Section below "**Different types of data objects**", we
will use different types of objects that can store different classes.

## R Operators

### Mathematical operators

Operators are symbols that perform mathematical calculations and other 
operations.

In basic arithmetic we use + (addition), - (subtraction), 
× (multiplication), ÷ (division).

The × and ÷ symbols are not on computer keyboards, so we 
substitute `*` for ×, and `/` for ÷.

```{r}
2 + 2
```

```{r}
4 - 2
```

```{r}
11 * 12
```

```{r}
22/7
```

There are numerous other operators in R as well, such as `<-`, `:`, `()`, `[]`, `^`, `&`, `|`, `==`, `%in%`, . . .

We also use simple mathematical functions like `log()`, `log10()`,
`sqrt()`, and so on.  
Remember that function names are always followed by parentheses `()`,
and within the parentheses we put the *arguments*, the input and 
information that the function needs.

The following chunks of code should help you understand how these
operators and simple functions work. 
*If you have any questions, please ask!*

#### The : operator

```{r}
# using the `:` operator
x <- 1:10
x
y <- 42:1
y
```

(Remember that we created objects called `x` and `y` before, so we 
have just replaced them with new values.)

#### Indexing using [ ] -- the square brackets operator

You will notice that each line of output above is preceded by `[1]` 
or another value.

The values in square brackets `[]` are the *index* of the value at the
beginning of each line of output. We can use the index to extract a
subset of values in an object (*make sure you understand* what is
happening here):

```{r}
y[3]
y[33:42]
```

#### Parentheses operator

```{r}
# using parentheses to control the order of operations
(8+1)/3
8+(1/3)
```

#### 'To the power of' caret (^) operator

```{r}
# using the `^` operator (raise to the power of)
10^3
```

We can calculate roots of numbers by raising to a reciprocal power. For example for the cube root of 8:

```{r}
# fractional powers give roots
8^(1/3)
```

#### The == operator

```{r}
# using the `==` operator (exactly equal to)

# we don't use `=` as this substitutes for `<-`, 
# and `=` is also used to set values for function arguments
a <- 5
a == 2+3
a == 6
```

#### The logical AND '&' operator

```{r}
# using the `&` operator (logical AND)
a <- 5
a == 2+3 & a == 10-5
a == 2+3 & a == 13-7
a == 4+3 & a == 10-5
```
#### The logical OR '|' operator

```{r}
# using the `|` operator (logical OR)
a <- 5
a == 2+3 | a == 10-5
a == 5 | a == 99
a == 7 | a == 3
```

### Simple mathematical functions

```{r}
# using simple mathematical functions
sqrt(64)
log(10)    # natural log (to base e ≈ 2.718282)
log10(10)  # log to base 10
```

### The %in% operator

The `%in%` operator checks if there is a match between one value or
vector, and a second vector or matrix.

```{r}
# in other words, the operator `%in%` asks "is a subset of"

j <- 2*(1:10)  # just even numbers up to 20
j
12 %in% j
17 %in% j
```

```{r}
# %in% works with character values too (note case-sensitivity)
z
"Helium" %in% z
"helium" %in% z
"Mars" %in% z
```

```{r}
# %in% works with multiple values (i.e. vectors) too
c("Tuesday", "Lithium", "Venus") %in% z
```

## Different types of data objects

### Matrices

We have just been using one-dimensional *vectors* so far.

Two-dimensional objects include *matrices*, *lists* and *data frames*.

```{r}
mymatrix <- matrix(101:109, ncol=3, byrow=TRUE)
mymatrix
```

Note that we included more than one argument in the `matrix()` 
function to make our 3×3 matrix:

| Argument  | Meaning                                      |
|:----------|:---------------------------------------------|
| `101:109` | Sequence of numbers 101, 102, ... 109        |
| `ncol`    | Number of columns                            |
| `byrow`   | If TRUE, put values in the matrix row by row |

A matrix can also be indexed but, since it is a two-dimensional array,
our index needs to have two values `[row,column]`, one of which can be
blank. Try the following and see if you can understand what is
happening:

```{r}
mymatrix[1,1]
mymatrix[1,3]
mymatrix[3,]
mymatrix[,1]
```

### Data frames

The final type of data object in this session is the *data frame*.  
This will be the most common type of data object we use in R, as it 
can store several different types of information.

```{r}
# making a data frame by directly inputting values
brightstars <- data.frame(Star=c("Alpha Centauri","Betelgeuse","Canopus","Sirius"),
                   Magnitude=c(-0.1, 0.42, -0.74, -1.46),
                   Light.Years=c(4.395, 427.5, 312.7, 8.611),
                   Planets=c(TRUE,FALSE,FALSE,FALSE))
brightstars
```

We can get useful information about a data frame with the `str()` 
(*structure*) or `summary()` functions:

```{r}
# `str()` (structure) function
str(brightstars)
```

```{r}
# summary() function
summary(brightstars)
```

In a data frame, it's common for the columns to be named (column
numbers are still present but invisible). We can use names or numbers
to index a data frame, remembering that it has two dimensions. Each
column behaves like a vector.

```{r}
brightstars[2,3]
brightstars[,"Star"]
brightstars[2:3,"Magnitude"]
```

NOTE -- see also how `%in%` works with data frames:

```{r}
"Sirius" %in% brightstars       # brightstars is a data frame
"Sirius" %in% brightstars$Star  # brightstars$Star is a vector
```

### Reading data from a file

More commonly we read data into an R data frame from a file. You will
need to download the data file `cities.csv` from LMS and save it into
your R project working directory (folder). (The data are for different
areas of land use in global cities).

You can also download the file from <https://github.com/Ratey-AtUWA/Learn-R-web/raw/refs/heads/main/cities_Hu_etal_2021.csv>. Save it in your R project working directory (folder) as `cities.csv` (use right-click » Save link as).

```{r paged.print=FALSE}
cities <- read.csv("cities.csv", stringsAsFactors = TRUE)
cities
```

When we display the object `cities` by typing its name, the values at 
left are the *row numbers*. 

Let's check the *structure* of the data frame `cities` using `str()`:

```{r}
str(cities)
```

The effect of the argument `stringsAsFactors = TRUE` above was to
convert all "string" (i.e. text) information into *factors*, which
is how R stores categorical information. This makes sense for the
columns `Type`, `Global`, and `Region`, which classify each city into
two of more categories. The column `City`, however, is just unique
names, so doesn't need to be a factor (although this won't affect most
data analysis we might do). We can convert the`City` column back to
text (*character*):

Note that we can define a data frame column in two ways, e.g. for
the `City` column: `cities$City` is equivalent to `cities[,"City"]`  
(both are also equivalent to `cities[,1]`). 

```{r}
cities$City <- as.character(cities$City)
str(cities)
```

The function `as.character()` is an example of *coercion* in R; we try
to force an object into another class. In this case we have the column
object `City` within the data frame object `cities`, which is
initially of class `Factor`, and `as.character()` converts it to class
`chr` (character).

[A *list* is a more general type of data object than a data frame --
not all columns need to be the same length. The output of some R
functions can be a list.]

### Writing data to files

#### Saving your workspace

If we have a lot of objects in our workspace and we don’t want to have
to re-create them each time we use R, we can save our workspace. This
is especially useful if we have started an *R project* (which you
should have done already!).

To save the workspace there are two options:

1. In the RStudio `Environment` tab (top-right of the screen,
usually), click the 'Save workspace as' icon (💾). You should name the
file *without a prefix*, just the extension (i.e. nothing before the
period `.` character. So, in the Save As dialog box, the file name
should be `.RData`.

2. Run this code:
```{r}
save.image()
```

|         By default the `save.image()` function saves your workspace 
|         as `.Rdata`, which is what we want. With R Projects, any 
|         file called `.RData` will be loaded automatically when you 
|         open the project! This is very convenient.

#### Saving individual data objects

Sometimes we may want to save a copy of a data frame we’ve been
working on, especially if we have made substantial changes (e.g.
calculated new columns, which we will learn how to do later). We can
save in various formats, but we recommend the simple .csv format. To
save a `.csv` file, we use the function `write.csv()`:

```{r}
write.csv(cities, file = "cities_edited.csv")
```

You should now be able to see the file `cities_edited.csv` in the
Files pane at the lower right of the RStudio window. (The file name
can be anything you like, but it makes sense to keep the `.csv`
suffix.)

#### Saving our markdown file(s)

We should always save any R markdown files that we create by clicking
the 💾 icon, typing Ctrl-S (⌘-S on Mac), or clicking File/Save in the
menu. By default, R markdown files will be saved into our project
folder (i.e. working directory), and will have the extension `.Rmd`
after the name.

#### Saving our markdown as a document

One of the main reasons we write R code in R Markdown is that we can
convert or *'knit'* our files to document formats. At the top of this
RStudio pane there is a `Knit ▼` button. Clicking this will save a
copy of our R markdown file as the default format (html) into our
project folder, or by clicking directly on the `▼` we can select
another format. Try it!  
(Note that `Knit to PDF` will not work unless we have a Latex app 
installed).

Knitting `.Rmd` files into documents is one of the coolest things
about R Markdown, and an excellent way to prepare *reports*, as we
have all the data analysis for the report included with the report
text. With advanced use of R markdown we can hide the R code and only
include the output we want!

## Getting Help

RStudio has an excellent help system. For help on any function, try
running code like `?function.name`  
(this is the equivalent of `help(function.name))`. Try it:

```{r}
?write.csv
```

The help topic will open in the `Help` pane at the right of the 
RStudio window.

More general help on many R topics can be obtained by running the
following code (try this too):

```{r}
help.start()
```

You will get a short message in the output box above, and a page with
links to a number of R help topics will open in the Help pane at the
right of the RStudio window.

You may also notice, when typing **code**, that RStudio has
*predictive text* that can help you get what you need. A pop-up should also appear, which shows the general form of the function you’re 
typing (i.e. the various inputs and options).

Finally, there is another useful advantage to using R Markdown in
RStudio – when typing **text** there is a *spell-checker*, shown as wavy
underlines beneath misspelled words. Right-clicking on the misspelled
word will show a list of correct options, just like a word processor.

🔚
